### 2015-01-03 paper/BMC_Bioinformatics_2015/Fig_MelnikovDiffPrefs/Fig_MelnikovDiffPrefs.tex
Added manuscript draft for BMC Bioinformatics

\documentclass[12pt,letterpaper]{article}
\usepackage{mathptmx}
\usepackage{graphicx}
\usepackage{setspace}
\usepackage{rotating}
\usepackage[export]{adjustbox}
\pagestyle{empty}
\usepackage[paperheight=5.5in,paperwidth=6.4in,top=0.05in,bottom=0.05in,right=0.05in,left=0.05in]{geometry}

\begin{document}
\begin{figure}

{\bf \Large A} \\
\includegraphics[clip, trim=0 0 0 0.04in,valign=t,width=5.2in]{../../../examples/Melnikov_et_al_Tn5/infer_diffprefs_on_simulated_data/inferred_diffprefs_logoplot.pdf}
\vspace{0.15in}

{\bf \Large B} \\
\includegraphics[clip, trim=0.12in 0 0 0.05in,valign=t,width=6.3in]{../../../examples/Melnikov_et_al_Tn5/infer_diffprefs_on_simulated_data/correlations.pdf}

\end{figure}
\end{document}

### 2015-01-03 paper/BMC_Bioinformatics_2015/Fig_MelnikovPrefs/Fig_MelnikovPrefs.tex
Added manuscript draft for BMC Bioinformatics

\documentclass[12pt,letterpaper]{article}
\usepackage{mathptmx}
\usepackage{graphicx}
\usepackage{setspace}
\usepackage{rotating}
\usepackage[export]{adjustbox}
\pagestyle{empty}
\usepackage[paperheight=4.1in,paperwidth=6.73in,top=0.05in,bottom=0.05in,right=0.05in,left=0.05in]{geometry}

\begin{document}
\begin{figure}
\begin{minipage}[t][][t]{5in}
{\bf \Large A} \\
\includegraphics[clip, trim=0 0 0 0.04in,valign=t,width=5in]{../../../examples/Melnikov_et_al_Tn5/prefs_logoplot.pdf}
\end{minipage}
\hspace{0.05in}
\begin{minipage}[t][][t]{1.5in}
{\bf \Large B} \\
\vspace{-0.2in}
\includegraphics[valign=t,width=1.5in]{../../../examples/Melnikov_et_al_Tn5/replicate_pref_corr.pdf}
\end{minipage}
\end{figure}
\end{document}

### 2015-01-03 paper/BMC_Bioinformatics_2015/Fig_ThyagarajanPrefs/Fig_ThyagarajanPrefs.tex
Added manuscript draft for BMC Bioinformatics

\documentclass[12pt,letterpaper]{article}
\usepackage{mathptmx}
\usepackage{graphicx}
\usepackage{setspace}
\usepackage{rotating}
\usepackage[export]{adjustbox}
\pagestyle{empty}
\usepackage[paperheight=7.6in,paperwidth=6.3in,top=0.05in,bottom=0.05in,right=0.05in,left=0.05in]{geometry}

\begin{document}
\begin{figure}

{\bf \Large A} \\
\includegraphics[clip, trim=0 0 0 0.04in,valign=t,width=6.3in]{../../../examples/ThyagarajanBloom2014_WSN_HA/prefs_logoplot.pdf}
\vspace{0.07in}

{\bf \Large B} \\
\vspace{-0.2in}
\includegraphics[valign=t,width=1.5in]{../../../examples/ThyagarajanBloom2014_WSN_HA/correlations/corr_1_vs_2.pdf}\hspace{0.2in}
\includegraphics[valign=t,width=1.5in]{../../../examples/ThyagarajanBloom2014_WSN_HA/correlations/corr_1_vs_3.pdf}\hspace{0.2in}
\includegraphics[valign=t,width=1.5in]{../../../examples/ThyagarajanBloom2014_WSN_HA/correlations/corr_2_vs_3.pdf}

\end{figure}
\end{document}

### 2015-01-03 paper/BMC_Bioinformatics_2015/bmc_bioinformatics_article.tex
Added manuscript draft for BMC Bioinformatics

%% BioMed_Central_Tex_Template_v1.06
%%                                      %
%  bmc_article.tex            ver: 1.06 %
%                                       %

%%IMPORTANT: do not delete the first line of this template
%%It must be present to enable the BMC Submission system to
%%recognise this template!!

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                     %%
%%  LaTeX template for BioMed Central  %%
%%     journal article submissions     %%
%%                                     %%
%%          <8 June 2012>              %%
%%                                     %%
%%                                     %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                                                 %%
%% For instructions on how to fill out this Tex template           %%
%% document please refer to Readme.html and the instructions for   %%
%% authors page on the biomed central website                      %%
%% http://www.biomedcentral.com/info/authors/                      %%
%%                                                                 %%
%% Please do not use \input{...} to include other tex files.       %%
%% Submit your LaTeX manuscript as one .tex document.              %%
%%                                                                 %%
%% All additional figures and files should be attached             %%
%% separately and not embedded in the \TeX\ document itself.       %%
%%                                                                 %%
%% BioMed Central currently use the MikTex distribution of         %%
%% TeX for Windows) of TeX and LaTeX.  This is available from      %%
%% http://www.miktex.org                                           %%
%%                                                                 %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%% additional documentclass options:
%  [doublespacing]
%  [linenumbers]   - put the line numbers on margins

%%% loading packages, author definitions

\documentclass[twocolumn]{bmcart}% uncomment this for twocolumn layout and comment line below
%\documentclass{bmcart}

%%% Load packages
\usepackage{amsthm,amsmath}
%\RequirePackage{natbib}
\RequirePackage{hyperref}
\usepackage[utf8]{inputenc} %unicode support
\usepackage{graphicx}
\usepackage{cprotect}
\usepackage{listings}
\lstset{
basicstyle=\footnotesize\ttfamily,
columns=flexible,
breaklines=true,
breakatwhitespace=true,
aboveskip=3pt,
belowskip=3pt,
}
\allowdisplaybreaks[2]
%\usepackage[applemac]{inputenc} %applemac support if unicode package fails
%\usepackage[latin1]{inputenc} %UNIX support if unicode package fails

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                             %%
%%  If you wish to display your graphics for   %%
%%  your own use using includegraphic or       %%
%%  includegraphics, then comment out the      %%
%%  following two lines of code.               %%
%%  NB: These line *must* be included when     %%
%%  submitting to BMC.                         %%
%%  All figure files must be submitted as      %%
%%  separate graphics through the BMC          %%
%%  submission process, not included in the    %%
%%  submitted article.                         %%
%%                                             %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%\def\includegraphic{}
%\def\includegraphics{}

%%% Put your definitions there:
\startlocaldefs
\endlocaldefs

%%% Begin ...
\begin{document}

%%% Start of article front matter
\begin{frontmatter}

\begin{fmbox}
\dochead{Software}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Enter the title of your article here     %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\title{Software for the analysis and visualization of deep mutational scanning data}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Enter the authors here                   %%
%%                                          %%
%% Specify information, if available,       %%
%% in the form:                             %%
%%   <key>={<id1>,<id2>}                    %%
%%   <key>=                                 %%
%% Comment or delete the keys which are     %%
%% not used. Repeat \author command as much %%
%% as required.                             %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\author[
   addressref={aff1},                   % id's of addresses, e.g. {aff1,aff2}
   corref={aff1},                       % id of corresponding address, if any
   email={jbloom@fredhutch.org}   % email address
]{\inits{JD}\fnm{Jesse D.} \snm{Bloom}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Enter the authors' addresses here        %%
%%                                          %%
%% Repeat \address commands as much as      %%
%% required.                                %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\address[id=aff1]{%                           % unique id
  \orgname{Division of Basic Sciences and Computational Biology Program, Fred Hutchinson Cancer Research Center}, % university, etc
  \street{1100 Fairview Ave N},                     %
  \postcode{98109}                                % post or zip code
  \city{Seattle, WA},                              % city
  \cny{USA}                                    % country
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Enter short notes here                   %%
%%                                          %%
%% Short notes will be after addresses      %%
%% on first page.                           %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{artnotes}
%\note{Sample of title note}     % note to the article
%\note[id=n1]{Equal contributor} % note, connected to author
\end{artnotes}

\end{fmbox}% comment this for two column layout

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% The Abstract begins here                 %%
%%                                          %%
%% Please refer to the Instructions for     %%
%% authors on http://www.biomedcentral.com  %%
%% and include the section headings         %%
%% accordingly for your article type.       %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{abstractbox}

\begin{abstract} % abstract
\parttitle{Background} %if any
Deep mutational scanning is a technique to estimate the impacts of mutations on a gene by using deep sequencing to count mutations in a library of variants before and after imposing a functional selection. The impacts of mutations must be inferred from changes in their counts after selection.

\parttitle{Results} %if any
I describe a software package, \texttt{dms\_tools}, to infer the impacts of mutations from deep mutational scanning data using a likelihood-based treatment of the mutation counts. I show that \texttt{dms\_tools} yields more accurate inferences on simulated data than the widely used but statistically biased approach of calculating ratios of counts pre- and post-selection. Using \texttt{dms\_tools}, one can infer the preference of each site for each amino acid given a single selection pressure, or assess the extent to which these preferences change under different selection pressures. The preferences and their changes can be intuitively visualized with sequence-logo-style plots created using an extension to \texttt{weblogo}.

\parttitle{Conclusions}
\texttt{dms\_tools} implements a statistically principled approach for the analysis and subsequent visualization of deep mutational scanning data.
\end{abstract}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% The keywords begin here                  %%
%%                                          %%
%% Put each keyword in separate \kwd{}.     %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{keyword}
\kwd{deep mutational scanning}
\kwd{sequence logo}
\kwd{amino-acid preferences}
\end{keyword}

% MSC classifications codes, if any
%\begin{keyword}[class=AMS]
%\kwd[Primary ]{}
%\kwd{}
%\kwd[; secondary ]{}
%\end{keyword}

\end{abstractbox}
%
%\end{fmbox}% uncomment this for twcolumn layout

\end{frontmatter}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% The Main Body begins here                %%
%%                                          %%
%% Please refer to the instructions for     %%
%% authors on:                              %%
%% http://www.biomedcentral.com/info/authors%%
%% and include the section headings         %%
%% accordingly for your article type.       %%
%%                                          %%
%% See the Results and Discussion section   %%
%% for details on how to create sub-sections%%
%%                                          %%
%% use \cite{...} to cite references        %%
%%  \cite{koon} and                         %%
%%  \cite{oreg,khar,zvai,xjon,schn,pond}    %%
%%  \nocite{smith,marg,hunn,advi,koha,mouse}%%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%% start of article main body
% <put your article body there>

%%%%%%%%%%%%%%%%
%% Background %%
%%

\section*{Background}
Deep mutational scanning is a high-throughput experimental technique to assess the impacts of mutations on a protein-coding gene~\cite{fowler2014deep}. Figure \ref{fig:DeepMutationalScan} shows a schematic of deep mutational scanning. A gene is mutagenized, and the library of resulting variants is introduced into cells or viruses, which are then subjected to an experimental selection that enriches for functional variants and depletes non-functional ones. Deep sequencing of the variants pre- and post-selection provides information about the functional impacts of mutations. Since the original description of deep mutational scanning by Fowler \textit{et al}~\cite{fowler2010high}, the technique has been applied to a wide range of genes~\cite{traxlmayr2012construction,mclaughlin2012spatial,starita2013activity,melamed2013deep,roscoe2013analyses,firnberg2014comprehensive,bloom2014experimentally,melnikov2014comprehensive,thyagarajan2014inherent,wu2014high,wu2014ns,olson2014comprehensive}, both to measure mutational tolerance given a single selection pressure as in Figure \ref{fig:DeepMutationalScan}A, or to identify mutations that have different effects under alternative selections as in Figure \ref{fig:DeepMutationalScan}B. New techniques to create comprehensive codon-mutant libraries of genes make it possible to profile all single amino-acid mutations~\cite{firnberg2012,jain2013rapid,firnberg2014comprehensive,bloom2014experimentally,melnikov2014comprehensive}, while new techniques for \textit{in vivo} mutagenesis enable deep mutational scanning to be applied across the biological spectrum from viruses and bacteria to human cells~\cite{findlay2014saturation}.

A key component of deep mutational scanning is analysis of the data: First, raw reads from the deep sequencing must be processed to count mutations pre- and post-selection. Next, the biological effects of mutations must be inferred from these counts. The first task of processing the reads is idiosyncratic to the specific sequencing strategy used. But the second task of inferring mutational effects from sequencing counts is amenable to more general algorithms. However, only a few such algorithms have been described~\cite{fowler2011enrich,bank2014bayesian}. Here I present user-friendly software, \texttt{dms\_tools}, that infers mutational effects from sequencing counts. Before describing the algorithms implemented in \texttt{dms\_tools} and illustrating its use on existing and simulated data, I first discuss issues associated with inferring mutational effects from sequencing counts.

\subsection*{The nature of deep mutational scanning data.}
The data consist of counts of variants pre- and post-selection. The approach presented here treats each site in the gene separately, ignoring epistatic coupling among mutations. This aspect of the approach should not be construed as a suggestion that interactions among mutations are unimportant; indeed, several studies have used deep mutational scanning to examine pairwise epistasis~\cite{olson2014comprehensive,araya2012fundamental,bank2015systematic}, and techniques have been described to obtain linkage between distant sites~\cite{hiatt2010parallel,wu2014hiv}. However, the exploding combinatorics of multiple mutations (a 500-residue protein has only $19 \times 500 \approx 10^4$ single mutants, but $19^2 \times \frac{500 \times 499}{2} \approx 4 \times 10^7$  double mutants and $19^3 \times \frac{500!}{497! \times 3!} \approx 10^{11}$ triple mutants) make it currently plausible to comprehensively characterize only single mutations to all but the shortest genes. Therefore, treating sites independently is not a major limitation for most current datasets; eventually the approach here might be extended to include coupling among mutations.

The data for each site $r$ is characterized by the sequencing \emph{depth} (total number of counts); let $N_r^{\rm{pre}}$, $N_r^{\rm{post}}$, $N_r^{s1}$, and $N_r^{s2}$ denote the depth at $r$ for each of the four libraries in Figure \ref{fig:DeepMutationalScan}. Typical depths for current experiments are $N \sim 10^6$. Denote the counts of character $x$ (characters might be nucleotides, amino acids, or codons) at $r$ as $n_{r,x}^{\rm{pre}}$, $n_{r,x}^{\rm{post}}$, $n_{r,x}^{s1}$, and $n_{r,x}^{s2}$. The values of $n_{r,x}$ for characters $x$ that differ from the wildtype identity $\operatorname{wt}\left(r\right)$ depend on both the depth $N$ and the average per-site mutation rate. Typical experiments introduce about one mutation per gene, so $\overline{\mu} \sim 1/L$ where $L$ is the length of the gene. Therefore, if a 500-codon gene is sequenced at depth $N \sim 10^6$, we expect $N \overline{\mu} \sim 2000$ counts of non-wildtype codons at each site. Since there are 63 mutant codons, the average pre-selection counts for a mutation to a specific $x \ne \operatorname{wt}\left(r\right)$ will be $n_{r,x}^{\rm{pre}} \sim 30$, with counts for most mutations deviating from this average due to biases in creation of the mutant library and randomness in which molecules are sequenced. Counts in the post-selection libraries will further deviate from this average due to selection. Therefore, even at depths $N \sim 10^6$, the actual counts of most mutations will be quite modest.

\subsection*{The challenge of statistical inference from finite counts.}
The goal is to estimate the effects of mutations from changes in their counts after selection. Let $\mu_{r,x}$, $f_{r,x}$, $f_{r,x}^{s1}$, and $f_{r,x}^{s2}$ denote the \emph{true} frequencies of $x$ at $r$ that would be observed for the four libraries in Figure \ref{fig:DeepMutationalScan} if we sampled at infinite depth in both the actual experiment and the sequencing. If we knew these frequencies, we could calculate parameters that reflect the effects of mutations. One parameter that characterizes the effect of mutating $r$ from $\operatorname{wt}\left(r\right)$ to $x$ for the experiment in Figure \ref{fig:DeepMutationalScan}A is the \emph{enrichment ratio}
\begin{equation}
\label{eq:enrichment}
\phi_{r,x} = \frac{f_{r,x} / f_{r,\operatorname{wt}\left(r\right)}}{\mu_{r,x} / \mu_{r,\operatorname{wt}\left(r\right)}}.
\end{equation}
Beneficial mutations have $\phi_{r,x} > 1$, while deleterious ones have $\phi_{r,x} < 1$. A related parameter is the \emph{preference} $\pi_{r,x}$ of $r$ for $x$. At each site, the preferences are simply the enrichment ratios rescaled to sum to one:
\begin{equation}
\label{eq:preference}
\pi_{r,x} = \frac{\phi_{r,x}}{\sum_y \phi_{r,y}} = \frac{f_{r,x} / \mu_{r,x}}{\sum_y f_{r,y} / \mu_{r,y}},
\end{equation}
or equivalently
\begin{equation}
\label{eq:preference2}
f_{r,x} = \frac{\pi_{r,x} \times \mu_{r,x}}{\sum_y \pi_{r,y} \times \mu_{r,y}}.
\end{equation}
The preferences can be intuitively visualized (Figure \ref{fig:DeepMutationalScan}A) and interpreted as the equilibrium frequencies in substitution models for gene evolution~\cite{bloom2014experimentally,bloom2014informed} (after accounting for uneven mutational rates~\cite{yampolsky2005exchangeability,stoltzfus2009climbing}).

Equations \ref{eq:enrichment} and \ref{eq:preference} are in terms of the true frequencies $\mu_{r,x}$, $f_{r,x}$, etc. But in practice, we only observe the counts in the finite sample of sequenced molecules. The computational challenge is to estimate the preferences (or enrichment ratios) from these counts.

The most naive approach is to simply substitute the counts for the frequencies, replacing Equation \ref{eq:enrichment} with
\begin{equation}
\label{eq:enrichmentcounts}
\phi_{r,x} = \frac{\frac{n_{r,x}^{\rm{post}} + \mathcal{P}}{n_{r,\operatorname{wt}\left(r\right)}^{\rm{post}} + \mathcal{P}}}{\frac{n_{r,x}^{\rm{pre}} + \mathcal{P}}{n_{r,\operatorname{wt}\left(r\right)}^{\rm{pre}} + \mathcal{P}}}
\end{equation}
where $\mathcal{P}$ (often chosen to be one) is a pseudocount added to each count to avoid ratios of zero or infinity.

However, Equation \ref{eq:enrichmentcounts} involves ratios of counts with values $\sim 10$ to 100 -- and as originally noted by Karl Pearson~\cite{pearson1896mathematical,pearson1910}, ratios estimated from finite counts are statistically biased, with the bias increasing as the magnitude of the counts decrease. This bias can propagate into subsequent analyses, since many statistical tests assume symmetric errors. The problems caused by biases in uncorrected ratios have been noted even in applications such as isotope-ratio mass spectrometry~\cite{ogliore2011} and fluorescent imaging~\cite{van2000mean}, where the counts usually far exceed those in deep mutational scanning.

Taking ratios also abrogates our ability to use the magnitude of the counts to assess our certainty about conclusions. For instance, imagine that at a fixed depth, the counts of a mutation increase from a pre-selection value of 5 to a post-selection value of 10. While this doubling suggests that the mutation might be beneficial, the small counts make us somewhat uncertain of this conclusion. But if the counts increased from 20 to 40 we would be substantially more certain, and if they increased from 100 to 200 we would be quite sure. So only by an explicit statistical treatment of the counts can we fully leverage the data.

Here I describe a software package, \texttt{dms\_tools}, that infers mutational effects in a Bayesian framework using a likelihood-based treatment of the counts. This software can be used to infer and visualize site-specific preferences from experiments like Figure \ref{fig:DeepMutationalScan}A, and to infer and visualize differences in preferences under alternative selections from experiments like Figure \ref{fig:DeepMutationalScan}B.

\section*{Implementation and Results}

\subsection*{Algorithm to infer site-specific preferences.}
\texttt{dms\_tools} uses a Bayesian approach to infer site-specific preferences from experiments like those in Figure \ref{fig:DeepMutationalScan}A. The algorithm calculates the likelihoods of the counts given the unknown preferences and mutation / error rates, placing plausible priors over these unknown parameters. The priors correspond to the assumption that all possible identities (e.g. amino acids) have equal preferences, and that the mutation and error rates for each site are equal to the overall average for the gene. MCMC is used to calculate the posterior probability of the preferences given the counts.

This algorithm a slight modification of that in the \textsl{Methods} of \cite{bloom2014experimentally}; here the algorithm is described anew to explain the implementation in \texttt{dms\_tools}.

\subsubsection*{Optional controls to quantify error rates.}
Some sequencing reads that report a mutation may actually reflect an error. Errors can be quantified by sequencing an unmutated gene, so that any counts at $r$ of $x \ne \operatorname{wt}\left(r\right)$ for this control reflect errors. In some cases (e.g. sequencing an RNA virus where the post-selection libraries must be reverse-transcribed), error rates for the pre- and post-selection libraries may differ and so be described by different controls. Let $N_r^{\textrm{errpre}}$ and $N_r^{\rm{errpost}}$ be the depth and $n_{r,x}^{\rm{errpre}}$ and $n_{r,x}^{\rm{errpost}}$ be the counts of $x$ in the pre-selection and post-selection error controls, respectively. Define $\epsilon_{r,x}$ and $\rho_{r,x}$ to be the true frequencies of errors at $r$ from $\operatorname{wt}\left(r\right)$ to $x$ in the pre- and post-selection controls, respectively.

\subsubsection*{Likelihoods of observing specific mutational counts.}
Define vectors of the counts and frequencies for all characters at each site $r$, i.e. $\mathbf{n_r^{\textbf{pre}}}= \left(\cdots, n_{r,x}^{\textrm{pre}}, \cdots\right)$, $\mathbf{n_r^{\textbf{post}}}= \left(\cdots, n_{r,x}^{\textrm{post}}, \cdots\right)$, $\boldsymbol{\mathbf{\mu_r}} = \left(\cdots, \mu_{r,x}, \cdots\right)$, $\boldsymbol{\mathbf{f_r}} = \left(\cdots, f_{r,x}, \cdots\right)$, etc. Also define $\boldsymbol{\mathbf{\pi_r}}= \left(\cdots, \pi_{r,x}, \cdots\right)$ of the preferences for each $r$, noting that Equation \ref{eq:preference2} implies $\boldsymbol{\mathbf{f_r}} = \frac{\boldsymbol{\mathbf{\mu_r}}\circ\boldsymbol{\mathbf{\pi_r}}}{\boldsymbol{\mathbf{\mu_r}}\cdot \boldsymbol{\mathbf{\pi_r}}}$ where $\circ$ is the Hadamard product.

The likelihoods of some specific set of counts are:
\begin{gather}
\begin{split} \Pr&\left(\mathbf{n_r^{\textbf{errpre}}} \mid N_r^{\textrm{errpre}}, \boldsymbol{\mathbf{\epsilon_r}}\right) = \\& \operatorname{Multi}\left(\mathbf{n_r^{\textbf{errpre}}}; N_r^{\textrm{errpre}}, \boldsymbol{\mathbf{\epsilon_r}}\right) \label{eq:pr_nrerrpre} \end{split} \\
\begin{split} \Pr&\left(\mathbf{n_r^{\textbf{errpost}}} \mid N_r^{\textrm{errpost}}, \boldsymbol{\mathbf{\rho_r}}\right) = \\ & \operatorname{Multi}\left(\mathbf{n_r^{\textbf{errpost}}}; N_r^{\textrm{errpost}}, \boldsymbol{\mathbf{\rho_r}}\right) \label{eq:pr_nrerrpost} \end{split} \\
\begin{split} \Pr&\left(\mathbf{n_r^{\textbf{pre}}}\mid N_r^{\textrm{pre}}, \boldsymbol{\mathbf{\mu_r}}, \boldsymbol{\mathbf{\epsilon_r}}\right) = \\ &\operatorname{Multi}\left(\mathbf{n_r^{\textbf{pre}}}; N_r^{\textrm{pre}}, \boldsymbol{\mathbf{\mu_r}}+ \boldsymbol{\mathbf{\epsilon_r}}- \boldsymbol{\mathbf{\delta_r}}\right) \label{eq:pr_nrpre} \end{split} \\
\begin{split} \Pr&\left(\mathbf{n_r^{\textbf{post}}} \mid N_r^{\textrm{post}}, \boldsymbol{\mathbf{\mu_r}}, \boldsymbol{\mathbf{\pi_r}}, \boldsymbol{\mathbf{\rho_r}}\right) = \\ &\operatorname{Multi}\left(\mathbf{n_r^{\textbf{post}}}; N_r^{\textrm{post}}, \frac{\boldsymbol{\mathbf{\mu_r}}\circ \boldsymbol{\mathbf{\pi_r}}}{\boldsymbol{\mathbf{\mu_r}}\cdot \boldsymbol{\mathbf{\pi_r}}} + \boldsymbol{\mathbf{\rho_r}}- \boldsymbol{\mathbf{\delta_r}}\right)\label{eq:pr_nrpost} \end{split}
\end{gather}
where $\operatorname{Multi}$ is the multinomial distribution, $\boldsymbol{\mathbf{\delta_r}} = \left(\cdots, \delta_{x, \operatorname{wt}\left(r\right)}, \cdots\right)$ is a vector with the element corresponding to $\operatorname{wt}\left(r\right)$ equal to one and all other elements zero ($\delta_{xy}$ is the Kronecker delta), and we have assumed that the probability that a site experiences both a mutation and an error is negligibly small.

\subsubsection*{Priors over the unknown parameters.}
We specify Dirichlet priors over the parameter vectors:
\begin{gather}
\Pr\left(\boldsymbol{\mathbf{\pi_r}}\right) = \operatorname{Dirichlet}\left(\boldsymbol{\mathbf{\pi_r}}; \alpha_{\pi} \times \mathbf{1}\right)   \label{eq:pr_pir} \\
\Pr\left(\boldsymbol{\mathbf{\mu_r}}\right) = \operatorname{Dirichlet}\left(\boldsymbol{\mathbf{\mu_r}}; \alpha_{\mu} \times \mathcal{N}_x \times \boldsymbol{\mathbf{a_{r,\mu}}}\right) \label{eq:pr_mur} \\
\Pr\left(\boldsymbol{\mathbf{\epsilon_r}}\right) = \operatorname{Dirichlet}\left(\boldsymbol{\mathbf{\epsilon_r}}; \alpha_{\epsilon} \times \mathcal{N}_x \times \boldsymbol{\mathbf{a_{r,\epsilon}}}\right) \label{eq:pr_epsilonr} \\
\Pr\left(\boldsymbol{\mathbf{\rho_r}}\right) = \operatorname{Dirichlet}\left(\boldsymbol{\mathbf{\rho_r}}; \alpha_{\rho} \times \mathcal{N}_x \times
 \boldsymbol{\mathbf{a_{r,\rho}}}\right) \label{eq:pr_rhor}
\end{gather}
where $\mathbf{1}$ is a vector of ones, $\mathcal{N}_x$ is the number of characters (64 for codons, 20 for amino acids, 4 for nucleotides), the $\alpha$'s are scalar concentration parameters $> 0$ (by default \texttt{dms\_tools} sets the $\alpha$'s to one). For codons, the error rate depends on the number of nucleotides being changed. The average error rates  $\overline{\epsilon_m}$ and $\overline{\rho_m}$ for $m$ nucleotide changes are estimated as
\begin{gather}
\overline{\epsilon_m} = \frac{1}{L}\sum\limits_r \frac{1}{N_r^{\textrm{errpre}}}\sum\limits_{x} n_{r,x}^{\textrm{errpre}}\times \delta_{m,D_{x,\operatorname{wt}\left(r\right)}} \label{eq:avgepsilonm} \\
\overline{\rho_m} = \frac{1}{L}\sum\limits_r \frac{1}{N_r^{\textrm{errpost}}}\sum\limits_{x} n_{r,x}^{\textrm{errpost}}\times \delta_{m,D_{x,\operatorname{wt}\left(r\right)}} \label{eq:avgrhom}
\end{gather}
where $D_{x,\operatorname{wt}\left(r\right)}$ is the number of nucleotide differences between $x$ and $\operatorname{wt}\left(r\right)$. Given these definitions, 
\begin{gather}
\boldsymbol{\mathbf{a_{r,\epsilon}}} = \left(\cdots, \sum\limits_m \frac{\overline{\epsilon}_{m}}{\mathcal{C}_m} \times \delta_{m,D_{x,\operatorname{wt}\left(r\right)}},\cdots\right) \label{eq:arepsilon} \\
\boldsymbol{\mathbf{a_{r,\rho}}} = \left(\cdots, \sum\limits_m \frac{\overline{\rho}_{m}}{\mathcal{C}_m} \times \delta_{m,D_{x,\operatorname{wt}\left(r\right)}}, \cdots\right) \label{eq:arrho} 
\end{gather}
where $\mathcal{C}_m$ is the number of mutant characters with $m$ changes relative to wildtype (for nucleotides $\mathcal{C}_0 = 1$ and $\mathcal{C}_1 = 3$; for codons $\mathcal{C}_0 = 1$, $\mathcal{C}_1 = 9$, $\mathcal{C}_2 = \mathcal{C}_3 = 27$).

Our prior assumption is that the mutagenesis introduces all mutant characters at equal frequency (this assumption is only plausible for codons if the mutagenesis is at the codon level as in~\cite{firnberg2012,jain2013rapid,firnberg2014comprehensive,bloom2014experimentally,melnikov2014comprehensive}). The per-site mutagenesis rate is estimated as
\begin{equation}
\label{eq:avgmu}
\overline{\mu} = \left(\frac{1}{L}\sum\limits_r \frac{1}{N_r^{\textrm{pre}}}\sum\limits_{x\ne \operatorname{wt}\left(r\right)} n_{r,x}^{\textrm{pre}}\right) - \sum\limits_{m \ge 1} \overline{\epsilon_m},
\end{equation}
so that
\begin{equation}
\boldsymbol{\mathbf{a_{r,\mu}}} = \left(\cdots, \frac{\overline{\mu}}{\mathcal{N}_x - 1} + \delta_{x,\operatorname{wt}\left(r\right)} \times \left[1 - \overline{\mu}\right] ,\cdots\right). \label{eq:armu}
\end{equation}

\subsubsection*{Character types: nucleotides, amino acids, or codons.}
\texttt{dms\_tools} allows four possibilities for the type of character for the counts and preferences. The first three possibilities are simple: the counts and preferences can both be for any of nucleotides, amino acids, or codons.

The fourth possibility is that the counts are for codons, but the preferences for amino acids. In this case, define a function mapping codons to amino acids, 
\begin{equation}
\label{eq:A}
\mathbf{A}\left(\mathbf{w}\right) = \left(\cdots, \sum\limits_x \delta_{a,\mathcal{A}\left(x\right)} \times w_x, \cdots\right)
\end{equation}
where $\mathbf{w}$ is a 64-element vector of codons $x$, $\mathbf{A}\left(\mathbf{w}\right)$ is a 20- or 21-element (depending on the treatment of stop codons) vector of amino acids $a$, and $\mathcal{A}\left(x\right)$ is the amino acid encoded by $x$. The prior over the preferences $\boldsymbol{\mathbf{\pi_r}}$ is still a symmetric Dirichlet (now only of length 20 or 21), but the priors for $\boldsymbol{\mathbf{\mu_r}}$, $\boldsymbol{\mathbf{\epsilon_r}}$, and $\boldsymbol{\mathbf{\rho_r}}$ are now Dirichlets parameterized by $\mathbf{A}\left(\boldsymbol{\mathbf{a_{r,\mu}}}\right)$, $\mathbf{A}\left(\boldsymbol{\mathbf{a_{r,\epsilon}}}\right)$ and $\mathbf{A}\left(\boldsymbol{\mathbf{a_{r,\rho}}}\right)$ rather than $\boldsymbol{\mathbf{a_{r,\mu}}}$, $\boldsymbol{\mathbf{a_{r,\epsilon}}}$, and $\boldsymbol{\mathbf{a_{r,\rho}}}$. The likelihoods are computed in terms of these transformed vectors after similarly transforming the counts to $\mathbf{A}\left(\mathbf{n_r^{\textbf{pre}}}\right)$, $\mathbf{A}\left(\mathbf{n_r^{\textbf{post}}}\right)$, $\mathbf{A}\left(\mathbf{n_r^{\textbf{errpre}}}\right)$, and $\mathbf{A}\left(\mathbf{n_r^{\textbf{errpost}}}\right)$.

\subsubsection*{Implementation.}
The program \texttt{dms\_inferprefs} in the \texttt{dms\_tools} package infers the preferences by using \texttt{pystan}~\cite{pystan-software:2014} to perform MCMC over the posterior defined by the product of the likelihoods and priors in Equations \ref{eq:pr_nrerrpre}, \ref{eq:pr_nrerrpost}, \ref{eq:pr_nrpre}, \ref{eq:pr_nrpost}, \ref{eq:pr_pir}, \ref{eq:pr_mur}, \ref{eq:pr_epsilonr}, and \ref{eq:pr_rhor}. The program runs four chains from different initial values, and checks for convergence by ensuring that the Gelman-Rubin statistic $\hat{R}$~\cite{gelman1992inference} is $<1.1$ and the effective sample size is $>100$; the number of MCMC iterations is increased until these convergence is achieved. The program \texttt{dms\_logoplot} in the \texttt{dms\_tools} package visualizes the posterior mean preferences via an extension to \texttt{weblogo}~\cite{crooks2004}.

\subsection*{Inferring preferences with \texttt{dms\_tools}.}
\subsubsection*{Application to actual datasets.}
Figures \ref{fig:MelnikovPrefs} and \ref{fig:ThyagarajanPrefs} illustrate application of \texttt{dms\_tools} to two existing datasets~\cite{melnikov2014comprehensive,thyagarajan2014inherent}. As these figures show, the \texttt{dms\_inferprefs} and \texttt{dms\_logoplot} programs infer and visualize the preferences with a few simple commands. Error controls can be included when available (they are not for Figure \ref{fig:MelnikovPrefs}, but are for Figure \ref{fig:ThyagarajanPrefs}). The runtime for the MCMC depends on the gene length and character type (codons are slowest, nucleotides fastest) -- but if the inference is parallelized across multiple CPUs (using the \texttt{--ncpus} option of \texttt{dms\_inferprefs}), the inference should take no more than a few hours. As shown in Figures \ref{fig:MelnikovPrefs} and \ref{fig:ThyagarajanPrefs}, the visualizations can overlay information about protein structure onto the preferences.

Figures \ref{fig:MelnikovPrefs} and \ref{fig:ThyagarajanPrefs} also illustrate use of \texttt{dms\_correlate} to assess the correlation between preferences inferred from different biological replicates~\cite{blainey2014points} of the experiment. The inclusion and analysis of such replicates provide the only sure way to fully assess the sources of noise associated with deep mutational scanning.

\subsubsection*{Testing on simulated data.}
To test the accuracy of preference-inference by \texttt{dms\_tools}, I simulated deep mutational scanning counts using the preferences in Figure \ref{fig:MelnikovPrefs}, both with and without errors quantified by appropriate controls. I then used \texttt{dms\_tools} to infer preferences from the simulated data, and also made similar inferences using simple ratio estimation (Equation \ref{eq:enrichmentcounts}). Figure \ref{fig:PrefCorrelations} shows the inferred preferences versus the actual values used to simulate the data. For simulations with mutation counts (quantified by the product $N\overline{\mu}$ of the depth and average per-site mutation rate) $\sim 1000$ to 2000, the inferences are quite accurate. Inferences made by \texttt{dms\_tools} are always more accurate than those obtained by simply taking ratios of mutation counts.

\subsection*{Algorithm to infer differential preferences.}
As shown in Figure \ref{fig:DeepMutationalScan}B, a useful extension to the experiment in Figure \ref{fig:DeepMutationalScan}A is to subject the functional variants to two different selection pressures to identify mutations favored by one pressure versus the other. While this experiment could in principle by analyzed by simply comparing the initial unselected mutants to the final variants after the two alternative selections, this approach is non-ideal. In experiments like Figure \ref{fig:DeepMutationalScan}A, many mutations are enriched or depleted to some extent by selection, since a large fraction of random mutations affect protein function~\cite{shortle1985,rennell1991,shafikhani1997,guo2004,bloom2005}. Therefore, the assumption that all mutations are equally tolerated (i.e. the preferences for a site are all equal, or the enrichment ratios are all one) is not a plausible null hypothesis for Figure \ref{fig:DeepMutationalScan}A, and so \texttt{dms\_tools} simply infers the preferences given a uniform Dirichlet prior rather than trying to pinpoint some subset of sites with unequal preferences.

But in Figure \ref{fig:DeepMutationalScan}B, the assumption that most mutations will be similarly selected is a plausible null hypothesis, since only a small subset of mutations will have markedly different effects under two related selections. Therefore, \texttt{dms\_tools} uses a different algorithm to infer the differential preferences under the two selections. This algorithm combines a prior that mildly favors differential preferences of zero with a likelihood-based analysis of the mutation counts to estimate posterior probabilities over the differential preferences.

\subsubsection*{Definition of the differential preferences.}
Given an experiment like Figure \ref{fig:DeepMutationalScan}B, let $f_{r,x}^{\textrm{start}}$ be the true frequency of character $x$ at site $r$ in the starting library (equivalent to the frequency $f_{r,x}^{\textrm{post}}$ in the figure), and let $f_{r,x}^{s1}$ and $f_{r,x}^{s2}$ be the frequencies after selections $s1$ and $s2$, respectively. The differential preference $\Delta\pi_{r,x}$ for $x$ at $r$ in $s2$ versus $s1$ is defined by:
\begin{gather}
\label{eq:controlpi}
f_{r,x}^{s1} = \frac{f_{r,x}^{\textrm{start}} \times \pi_{r,x}^{s1}}{\sum_y f_{r,y}^{\textrm{start}} \times \pi_{r,y}^{s1}} \\
\label{eq:deltapi}
f_{r,x}^{s2} = \frac{f_{r,x}^{\textrm{start}} \times \left(\pi_{r,x}^{s1} + \Delta\pi_{r,x}\right)}{\sum_y f_{r,y}^{\textrm{start}} \times \left(\pi_{r,y}^{s1} + \Delta\pi_{r,y}\right)}
\end{gather}
where $\pi_{r,x}^{s1}$ is the ``control preference'' and is treated as a nuisance parameter, and we have the constraints
\begin{gather}
\label{eq:diffprefsum}
0 = \sum\limits_x \Delta\pi_{r,x} \\
\label{eq:diffprefpluspref}
0 \le \pi_{r,x}^{s1} + \Delta\pi_{r,x} \le 1.
\end{gather}
If there is no difference in the effect of $x$ at $r$ between selections $s1$ and $s2$, then $\Delta\pi_{r,x} = 0$. If $x$ at $r$ is more preferred by $s2$ than $s1$, then $\Delta\pi_{r,x} > 0$; conversely if $x$ at $r$ is more preferred by $s1$ than $s2$, then $\Delta\pi_{r,x} < 0$.

\subsubsection*{Likelihoods of observing specific mutational counts.}
Define vectors of the counts as $\mathbf{n_r^{\textbf{start}}}= \left(\cdots, n_{r,x}^{\textrm{start}}, \cdots\right)$ for the post-selection functional variants that are subjected to the further selections, and as $\mathbf{n_r^{s1}}= \left(\cdots, n_{r,x}^{s1}, \cdots\right)$ and $\mathbf{n_r^{s2}}= \left(\cdots, n_{r,x}^{s2}, \cdots\right)$ for selections $s1$ and $s2$. We again allow an error control, but now assume that the same control applies to all three libraries (since they are all sequenced after a selection), and define the counts for this control as $\mathbf{n_r^{\textbf{err}}}= \left(\cdots, n_{r,x}^{\textrm{err}}, \cdots\right)$; the true error frequencies are denoted by $\xi_{r,x}$. Define vectors of the frequencies, errors, control preferences, and differential preferences: $\boldsymbol{\mathbf{f_r^{\textbf{start}}}}= \left(\cdots, f_{r,x}^{\textrm{start}}, \cdots\right)$, $\boldsymbol{\mathbf{f_r^{s1}}}= \left(\cdots, f_{r,x}^{s1}, \cdots\right)$, $\boldsymbol{\mathbf{f_r^{s2}}}= \left(\cdots, f_{r,x}^{s2}, \cdots\right)$, $\boldsymbol{\mathbf{\xi_r}}= \left(\cdots, \xi_{r,x}, \cdots\right)$, $\boldsymbol{\mathbf{\pi^{s1}_r}}= \left(\cdots, \pi^{s1}_{r,x}, \cdots\right)$, and $\boldsymbol{\mathbf{\Delta\pi_r}}= \left(\cdots, \Delta\pi_{r,x}, \cdots\right)$. Equations \ref{eq:controlpi} and \ref{eq:deltapi} imply
$\boldsymbol{\mathbf{f_r^{s1}}}= \frac{\boldsymbol{\mathbf{f_r^{\textbf{starat}}}}\circ \boldsymbol{\mathbf{\pi_r^{s1}}}}{\boldsymbol{\mathbf{f_r^{\textbf{start}}}}\cdot \boldsymbol{\mathbf{\pi_r^{s1}}}}$ and $\boldsymbol{\mathbf{f_r^{s2}}} = \frac{\boldsymbol{\mathbf{f_r^{\textbf{start}}}}\circ \left(\boldsymbol{\mathbf{\pi_r^{s1}}} + \boldsymbol{\mathbf{\Delta\pi_r}}\right)}{\boldsymbol{\mathbf{f_r^{\textbf{start}}}}\cdot \left(\boldsymbol{\mathbf{\pi_r^{s1}}} + \boldsymbol{\mathbf{\Delta\pi_r}}\right)}$.

The likelihoods of the counts will be multinomially distributed around the ``true'' frequencies, so 
\begin{gather}
\label{eq:pr_nrerr}
\Pr\left(\mathbf{n_r^{\textbf{err}}}\mid N_r^{\textrm{err}}, \boldsymbol{\mathbf{\xi_r}}\right) = \operatorname{Multi}\left(\mathbf{n_r^{\textbf{err}}}; N_r^{\textrm{err}}, \boldsymbol{\mathbf{\xi_r}}\right) \\
\label{eq:pr_nrstart}
\begin{split}
\Pr&\left(\mathbf{n_r^{\textbf{start}}}\mid N_r^{\textrm{start}}, \boldsymbol{\mathbf{f_r^{\textbf{start}}}}, \boldsymbol{\mathbf{\xi_r}}\right) = \\ & \operatorname{Multi}\left(\mathbf{n_r^{\textbf{start}}}; N_r^{\textrm{start}}, \boldsymbol{\mathbf{f_r^{\textbf{start}}}} + \boldsymbol{\mathbf{\xi_r}}- \boldsymbol{\mathbf{\delta_r}}\right) \end{split} \\
\label{eq:pr_nrs1}
\begin{split}
\Pr&\left(\mathbf{n_r^{s1}}\mid N_r^{s1}, \boldsymbol{\mathbf{f_r^{\textbf{start}}}}, \boldsymbol{\mathbf{\pi_r^{s1}}}, \boldsymbol{\mathbf{\xi_r}}\right) = \\ & \operatorname{Multi}\left(\mathbf{n_r^{s1}}; N_r^{s1}, \frac{\boldsymbol{\mathbf{f_r^{\textbf{start}}}}\circ \boldsymbol{\mathbf{\pi_r^{s1}}}}{\boldsymbol{\mathbf{f_r^{\textbf{start}}}}\cdot \boldsymbol{\mathbf{\pi_r^{s1}}}} + \boldsymbol{\mathbf{\xi_r}}- \boldsymbol{\mathbf{\delta_r}}\right) \end{split} \\
\label{eq:pr_nrs2}
\begin{split} \Pr&\left(\mathbf{n_r^{s2}}\mid N_r^{s2}, \boldsymbol{\mathbf{f_r^{\textbf{start}}}}, \boldsymbol{\mathbf{\pi_r^{s1}}}, \boldsymbol{\mathbf{\Delta\pi_r}}, \boldsymbol{\mathbf{\xi_r}}\right) = \\ & \operatorname{Multi}\left(\mathbf{n_r^{s2}}; N_r^{s2}, \frac{\boldsymbol{\mathbf{f_r^{\textbf{start}}}}\circ \left(\boldsymbol{\mathbf{\Delta\pi_r}} + \boldsymbol{\mathbf{\pi_r^{s1}}}\right)}{\boldsymbol{\mathbf{f_r^{\textbf{start}}}}\cdot \left(\boldsymbol{\mathbf{\Delta\pi_r}} + \boldsymbol{\mathbf{\pi_r^{s1}}}\right)} + \boldsymbol{\mathbf{\xi_r}}- \boldsymbol{\mathbf{\delta_r}}\right) \end{split}
\end{gather}
where we have assumed that the probability that a site experiences a mutation and an error in the same molecule is negligibly small.

\subsubsection*{Priors over the unknown parameters.}
We specify Dirichlet priors over the parameter vectors:
\begin{gather}
\label{eq:pr_controlpir}
\Pr\left(\boldsymbol{\mathbf{\pi_r^{s1}}}\right) = \operatorname{Dirichlet}\left(\boldsymbol{\mathbf{\pi_r^{s1}}}; \alpha_{\pi^{s1}} \times \mathbf{1}\right) \\
\label{eq:pr_xir}
\Pr\left(\boldsymbol{\mathbf{\xi_r}}\right) = \operatorname{Dirichlet}\left( \boldsymbol{\mathbf{\xi_r}}; \alpha_{\xi} \times \mathcal{N}_x \times \boldsymbol{\mathbf{a_{r,\xi}}}\right) \\
\label{eq:pr_frstart}
\begin{split}
\Pr&\left(\boldsymbol{\mathbf{f_r^{\textbf{start}}}}\right) = \\ &\operatorname{Dirichlet}\left( \boldsymbol{\mathbf{f_r^{\textbf{start}}}}; \alpha_{\textrm{start}} \times \mathcal{N}_x \times \boldsymbol{\mathbf{a_{r,\textbf{start}}}}\right) \end{split} \\
\label{eq:pr_deltapi}
\begin{split}
\Pr&\left(\boldsymbol{\mathbf{\Delta\pi_r}} \mid \boldsymbol{\mathbf{\pi_r^{s1}}}\right) = \\& \operatorname{Dirichlet}\left(\boldsymbol{\mathbf{\Delta\pi_r}}; \alpha_{\Delta\pi} \times \mathcal{N}_x \times \boldsymbol{\mathbf{\pi_r^{s1}}}\right) - \boldsymbol{\mathbf{\pi_r^{s1}}} \end{split}
\end{gather}
where \texttt{dms\_tools} by default sets all the scalar concentration parameters ($\alpha$'s) to one except $\alpha_{\Delta\pi}$, which is set to two corresponding to a weak expectation that the $\Delta\pi$ values are close to zero. The average error rate $\overline{\xi_m}$ for mutations with $m$ of nucleotide changes is 
\begin{equation}
\label{eq:avgxim}
\overline{\xi_m} = \frac{1}{L}\sum\limits_r \frac{1}{N_r^{\textrm{err}}}\sum\limits_{x} n_{r,x}^{\textrm{err}}\times \delta_{m,D_{x,\operatorname{wt}\left(r\right)}}, 
\end{equation}
and so
\begin{equation}
\label{eq:arxi}
\boldsymbol{\mathbf{a_{r,\xi}}} = \left(\cdots, \sum\limits_m \frac{\overline{\xi}_{m}}{\mathcal{C}_m} \times \delta_{m,D_{x,\operatorname{wt}\left(r\right)}},\cdots\right).
\end{equation}
Our prior assumption is that all mutations are at equal frequency in the starting library (this assumption is unlikely to be true if the starting library has already been subjected to some selection, but we lack a rationale for a more informative prior). The average mutation frequency in the starting library is
\begin{equation}
\label{eq:avgfstart}
\overline{f^{\textrm{start}}} = \left(\frac{1}{L}\sum\limits_r \frac{1}{N_r^{\textrm{start}}}\sum\limits_{x\ne \operatorname{wt}\left(r\right)} n_{r,x}^{\textrm{start}}\right) - \sum\limits_{m \ge 1} \overline{\xi_m},
\end{equation}
and so 
\begin{equation}
\label{eq:arstart}
\boldsymbol{\mathbf{a_{r,\textbf{start}}}}= \left(\cdots, \frac{\overline{f^{\textrm{start}}}}{\mathcal{N}_x - 1} + \delta_{x,\operatorname{wt}\left(r\right)} \times \left[1 - \overline{f^{\textrm{start}}}\right] ,\cdots\right).
\end{equation}

\subsubsection*{Implementation.}
The program \texttt{dms\_inferdiffprefs} in the \texttt{dms\_tools} package infers the differential preferences by performing MCMC over the posterior defined by the product of the likelihoods and priors in Equations \ref{eq:pr_nrerr}, \ref{eq:pr_nrstart}, \ref{eq:pr_nrs1}, \ref{eq:pr_nrs2}, \ref{eq:pr_controlpir}, \ref{eq:pr_xir}, \ref{eq:pr_frstart}, and \ref{eq:pr_deltapi}. The MCMC is performed as described for the preferences, and characters can again be any of nucleotides, amino acids, or codons. The program \texttt{dms\_logoplot} visualizes the posterior mean differential preferences via an extension to \texttt{weblogo}~\cite{crooks2004}.

\subsection*{Inferring differential preference with \texttt{dms\_tools}.}
To test the accuracy of differential preference inference by \texttt{dms\_tools}, I simulated an experiment like that in Figure \ref{fig:DeepMutationalScan}B with the starting counts based on Melnikov \textit{et al}'s actual deep mutational scanning data of a Tn5 transposon~\cite{melnikov2014comprehensive}. As show by Figure \ref{fig:MelnikovSimulatedDiffPrefs}, \texttt{dms\_inferdiffprefs} accurately infers the differential preferences at typical experimental depths. The results are easily visualized with \texttt{dms\_logoplot}. To provide a second illustration of differential preferences, Figure \ref{fig:WuDiffPrefs} shows an analysis of the data obtained by Wu \textit{et al} when they performed an experiment like that in Figure \ref{fig:DeepMutationalScan}B on nucleotide mutants of the influenza NS gene in the presence or absence of interferon treatment.

\section*{Conclusions}
\texttt{dms\_tools} is a freely available software package that uses a statistically principled approach to analyze deep mutational scanning data. This paper shows that \texttt{dms\_tools} accurately infers preferences and differential preferences from data simulated under realistic parameters. As the figures illustrate, \texttt{dms\_tools} can also be applied to actual data with a few simple commands. The intuitive visualizations created by \texttt{dms\_tools} assist in interpreting the results. As deep mutational scanning continues to proliferate as an experimental technique~\cite{fowler2014deep}, \texttt{dms\_tools} can be applied to analyze the data for purposes such as guiding protein engineering~\cite{traxlmayr2012construction,melnikov2014comprehensive}, understanding sequence-structure-function relationships~\cite{mclaughlin2012spatial,starita2013activity,roscoe2013analyses,olson2014comprehensive,araya2012fundamental}, informing the development of better evolutionary models for sequence analysis~\cite{bloom2014experimentally,bloom2014informed}, and probing the biology of viruses and cells~\cite{melamed2013deep,firnberg2014comprehensive,thyagarajan2014inherent,wu2014high,wu2014ns,findlay2014saturation}.

\section*{Availability and requirements}
\begin{itemize}
\item {\bf Project name:} \texttt{dms\_tools}
\item {\bf Project home page:}
\begin{itemize} 
\item Documentation and installation instructions: \url{http://jbloom.github.io/dms_tools/} 
\item Source code: \url{https://github.com/jbloom/dms_tools} 
\end{itemize}
\item {\bf Operating system(s):} Linux
\item {\bf Programming language:} \texttt{Python}
\item {\bf Other requirements:} \texttt{pystan}, \texttt{weblogo}
\item {\bf License:} GNU GPLv3
\item {\bf Restrictions to use by non-academics:} None
\end{itemize}

\section*{Data and code for figures in this paper.}
The data and computer code used to generate the figures are in the tagged version of the \texttt{dms\_tools} source code at \url{https://github.com/jbloom/dms_tools/tree/1.0} in \texttt{examples} subdirectory. The \texttt{LaTex} source for this paper is in the \texttt{paper} subdirectory.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Backmatter begins here                   %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{backmatter}

\section*{Competing interests}
  The author declares that he has no competing interests.

\section*{Author's contributions}
JDB designed the algorithms, wrote the software, performed the analyses, and wrote the paper.

\section*{Acknowledgements}
Thanks to Alec Heckert for assistance in testing \texttt{dms\_tools}, to Erick Matsen for the excellent suggestion to use \texttt{pystan} for MCMC, and to Nicholas Wu for providing the mutational counts data from \cite{wu2014ns}. This work was supported by the NIGMS of the NIH under grant R01GM102198.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                  The Bibliography                       %%
%%                                                         %%
%%  Bmc_mathpys.bst  will be used to                       %%
%%  create a .BBL file for submission.                     %%
%%  After submission of the .TEX file,                     %%
%%  you will be prompted to submit your .BBL file.         %%
%%                                                         %%
%%                                                         %%
%%  Note that the displayed Bibliography will not          %%
%%  necessarily be rendered by Latex exactly as specified  %%
%%  in the online Instructions for Authors.                %%
%%                                                         %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% if your bibliography is in bibtex format, use those commands:
\bibliographystyle{bmc-mathphys} % Style BST file
\bibliography{references}      % Bibliography file (usually '*.bib' )

% or include bibliography directly:
% \begin{thebibliography}
% \bibitem{b1}
% \end{thebibliography}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                               %%
%% Figures                       %%
%%                               %%
%% NB: this is for captions and  %%
%% Titles. All graphics must be  %%
%% submitted separately and NOT  %%
%% included in the Tex document  %%
%%                               %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%
%% Do not use \listoffigures as most will included as separate files

\clearpage
\onecolumn

\section*{Figures}

\begin{figure}[h!]
\centerline{\includegraphics[clip, trim=0in 0in 5.92in 0in, width=5.5in]{Fig_DeepMutationalScan/Fig_DeepMutationalScan.pdf}}
\caption{\label{fig:DeepMutationalScan}{\csentence{A deep mutational scanning experiment.} {\bf (A)} A gene is mutagenized to create a library that contains all single codon mutations. The mutant library is introduced into cells or viruses and subjected to a functional selection that enriches beneficial mutations and depletes deleterious ones. Deep sequencing is used to count mutations in a sample of the variants present pre- and post-selection. Using \texttt{dms\_tools}, the data can be analyzed to infer the ``preference'' of each site for each amino acid; in the visualization, letter heights are proportional to the preference for that amino acid. {\bf (B)} The experiment can be extended by subjecting the library of functional variants to two different selection pressures, and using deep sequencing to assess which variants are favored in one condition versus the other. Using \texttt{dms\_tools}, the data can be analyzed to infer the ``differential preference'' of each site for each amino acid in the alternative selection $s2$ versus the control selection $s1$; in the visualization, letter heights above or below the line are proportional to the differential preference for or against that amino acid.}}
\end{figure}

\begin{figure}[h!]
\centerline{\includegraphics[width=6.7in]{Fig_MelnikovPrefs/Fig_MelnikovPrefs.pdf}}
\cprotect \caption{\label{fig:MelnikovPrefs}\csentence{Site-specific preferences from deep mutational scanning of a Tn5 transposon.} Melnikov \textit{et al}~\cite{melnikov2014comprehensive} performed deep mutational scanning on a Tn5 transposon using kanamycin selection, and reported the counts of amino-acid mutations for two biological replicates of the experiment. Here I have used \texttt{dms\_tools} to infer the preferences. {\bf (A)} Visualization of the preferences averaged across the two replicates. {\bf (B)} Correlation between the preferences inferred from each of the two replicates. Given files containing the mutation counts, the plots can be generated as \texttt{logoplot.pdf} and \texttt{corr.pdf} with the following commands:
\begin{lstlisting} 
dms_inferprefs pre_counts_1.txt post_counts_1.txt prefs_1.txt --excludestop --ncpus -1 --chartype aa
dms_inferprefs pre_counts_2.txt post_counts_2.txt prefs_2.txt --excludestop --ncpus -1 --chartype aa
dms_correlate prefs_1.txt prefs_2.txt corr --name1 "replicate 1" --name2 "replicate 2" --corr_on_plot
dms_merge prefs.txt average prefs_1.txt prefs_2.txt
dms_logoplot prefs.txt logoplot.pdf --nperline 53 --overlay1 RSAs.txt RSA "relative solvent accessibility" --overlay2 SSs.txt SS "secondary structure"
\end{lstlisting}
}
\end{figure}

\begin{figure}[h!]
\centerline{\includegraphics[width=6.1in]{Fig_ThyagarajanPrefs/Fig_ThyagarajanPrefs.pdf}}
\cprotect \caption{\label{fig:ThyagarajanPrefs}\csentence{Site-specific preferences from deep mutational scanning of influenza hemagglutinin.} Thyagarajan and Bloom~\cite{thyagarajan2014inherent} performed deep mutational scanning on influenza hemagglutinin, and reported the counts of codon mutations for three biological replicates of the experiment. Here I have used \texttt{dms\_tools} to infer the preferences. {\bf (A)} Visualization of the preferences averaged across the three replicates. {\bf (B)} Correlations between the preferences from each pair of replicates. Given files containing the mutation counts, the plots can be generated as \texttt{logoplot.pdf}, \texttt{corr\_1\_2.pdf}, \texttt{corr\_1\_3.pdf}, and \texttt{corr\_2\_3.pdf} with the following commands:
\begin{lstlisting} 
dms_inferprefs mutDNA_1.txt mutvirus_1.txt prefs_1.txt --errpre DNA_1.txt --errpost virus_1.txt --ncpus -1
dms_inferprefs mutDNA_2.txt mutvirus_2.txt prefs_2.txt --errpre DNA_2.txt --errpost virus_2.txt --ncpus -1
dms_inferprefs mutDNA_3.txt mutvirus_3.txt prefs_3.txt --errpre DNA_3.txt --errpost virus_3.txt --ncpus -1
dms_correlate prefs_1.txt prefs_2.txt corr_1_2 --name1 "replicate 1" --name2 "replicate 2" --corr_on_plot
dms_correlate prefs_1.txt prefs_3.txt corr_1_3 --name1 "replicate 1" --name2 "replicate 3" --corr_on_plot
dms_correlate prefs_2.txt prefs_3.txt corr_2_3 --name1 "replicate 2" --name2 "replicate 3" --corr_on_plot
dms_merge prefs.txt average prefs_1.txt prefs_2.txt prefs_3.txt
dms_logoplot prefs.txt logoplot.pdf --nperline 71 --overlay1 RSAs.txt RSA "relative solvent accessibility" --overlay2 SSs.txt SS "secondary structure" --excludestop
\end{lstlisting}
}
\end{figure}

\begin{figure}[h!]
\centerline{\includegraphics[width=6.5in]{../../examples/Melnikov_et_al_Tn5/infer_prefs_on_simulated_data/correlations.pdf}}
\cprotect\caption{\label{fig:PrefCorrelations}\csentence{Accuracy of preference inference on simulated data.} Deep mutational scanning counts were simulated using the preferences in Figure \ref{fig:MelnikovPrefs}A and realistic parameters for mutation and error rates. The simulations were done {\bf (A)} without or {\bf (B)} with sequencing errors quantified by control libraries. Plots show the correlation between the actual and inferred preferences as a function of the product of the sequencing depth $N$ and the average per-site mutation rate $\overline{\mu}$; real experiments typically have $N\overline{\mu} \sim 1000$ to 2000 depending on the sequencing depth and gene length. Preferences are inferred using the full algorithm in \texttt{dms\_tools} (top panels) or by simply calculating ratios of counts (bottom panels) using Equation \ref{eq:enrichmentcounts} and its logical extension to include errors, both with a pseudocount of one. The \texttt{dms\_tools} inferences are more accurate than the simple ratio estimation, with both methods converging to the actual values with increasing $N\overline{\mu}$. Given files with the mutation counts, the plots in this figure can be generated as \texttt{prefs\_corr.pdf} and \texttt{ratio\_corr.pdf} with commands such as:
\begin{lstlisting}
dms_inferprefs pre.txt post.txt inferred_prefs.txt --ncpus -1
dms_inferprefs pre.text post.text ratio_prefs.txt --ratio_estimation 1
dms_correlate actual_prefs.txt inferred_prefs.txt prefs_corr --name1 "actual" --name2 "inferred" --corr_on_plot --r2
dms_correlate actual_prefs.txt ratio_prefs.txt ratio_corr --name1 "actual" --name2 "inferred" --corr_on_plot --r2
\end{lstlisting}
}
\end{figure}

\begin{figure}[h!]
\centerline{\includegraphics[width=6.5in]{Fig_MelnikovDiffPrefs/Fig_MelnikovDiffPrefs.pdf}}
\cprotect\caption{\label{fig:MelnikovSimulatedDiffPrefs}\csentence{Inference of differential preferences on simulated data.} To illustrate and test the inference of differential preferences, the experiment in Figure \ref{fig:DeepMutationalScan}B was simulated at the codon level starting with the post-selection library that yielded the preferences in Figure \ref{fig:MelnikovPrefs}. In the simulations, 20\% of sites had different preferences between the control and alternative selection. {\bf (A)}, \texttt{dms\_tools} was used to infer the differential preferences from the data simulated at $N = 10^7$, and the resulting inferences were visualized. The overlay bars indicate which sites had non-zero differential preferences in the simulation. {\bf (B)} The correlations between the inferred and actual differential preferences as a function of $N\overline{\mu}$ show that the inferred values converge to the true ones. Given files with the mutation counts, the plots in this figure can be generated as \texttt{logoplot.pdf} and \texttt{corr.pdf} with the following commands:
\begin{lstlisting}
dms_inferdiffprefs start.txt s1.txt s2.txt diffprefs.txt --ncpus -1 
dms_logoplot diffprefs.txt  logoplot.pdf --nperline 53 --overlay1 actually_nonzero.txt "$\ne 0$?" "Is actual differential preference non-zero?" --diffprefheight 0.45
dms_correlate actual_diffprefs.txt diffprefs.txt corr --name1 "actual" --name2 "inferred" --corr_on_plot --r2
\end{lstlisting}
}
\end{figure}

\begin{figure}[h!]
\centerline{\includegraphics[width=6.5in]{../../examples/Wu_et_al_NS1/diffprefs_logoplot.pdf}}
\cprotect\caption{\label{fig:WuDiffPrefs}\csentence{Differential preferences following selection of influenza NS1 in the presence or absence of interferon.} Wu \textit{et al}~\cite{wu2014ns} generated libraries of influenza viruses carrying nucleotide mutations in the NS segment. They passaged these viruses in the presence or absence of interferon pre-treatment. Here, \texttt{dms\_tools} was used to analyze and visualize the data to identify sites where different nucleotides are preferred in the presence versus the absence of interferon. Because the mutations were made at the nucleotide level, the data must also be analyzed at that level (unlike in Figures \ref{fig:MelnikovPrefs}, \ref{fig:ThyagarajanPrefs}, and \ref{fig:MelnikovSimulatedDiffPrefs}, where codon mutagenesis means that the data can be analyzed at the amino-acid level). The plot can be generated as \texttt{logoplot.pdf} with the following commands:
\begin{lstlisting}
dms_inferdiffprefs input.txt control.txt interferon.txt diffprefs.txt --ncpus -1 --chartype DNA
dms_logoplot diffprefs.txt logoplot.pdf --nperline 68 --diffprefheight 0.4
\end{lstlisting}
}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                               %%
%% Tables                        %%
%%                               %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%% Use of \listoftables is discouraged.
%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                               %%
%% Additional Files              %%
%%                               %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\end{backmatter}
\end{document}

### 2015-01-08 paper/BMC_Bioinformatics_2015/bmc_bioinformatics_article.tex
Made text changes to LaTex BMC Bioinformatics draft

@@
 Since the original description of deep mutational scanning by Fowler \textit{et al}~\cite{fowler2010high}, the technique has been applied to a wide range of genes~\cite{traxlmayr2012construction,mclaughlin2012spatial,starita2013activity,melamed2013deep,roscoe2013analyses,firnberg2014comprehensive,bloom2014experimentally,melnikov2014comprehensive,thyagarajan2014inherent,wu2014high,wu2014ns,olson2014comprehensive}, both to measure mutational tolerance given a single selection pressure as in Figure \ref{fig:DeepMutationalScan}A, or to identify mutations that have different effects under alternative selections as in Figure \ref{fig:DeepMutationalScan}B.
-New techniques to create comprehensive codon-mutant libraries of genes make it possible to profile all single amino-acid mutations~\cite{firnberg2012,jain2013rapid,firnberg2014comprehensive,bloom2014experimentally,melnikov2014comprehensive}, while new techniques for \textit{in vivo} mutagenesis enable deep mutational scanning to be applied across the biological spectrum from viruses and bacteria to human cells~\cite{findlay2014saturation}.
+New techniques to create comprehensive codon-mutant libraries of genes make it possible to profile all single amino-acid mutations~\cite{firnberg2012,jain2013rapid,firnberg2014comprehensive,bloom2014experimentally,melnikov2014comprehensive}, while new techniques for targeted mutagenesis of mammalian genomes enable deep mutational scanning to be applied across the biological spectrum from viruses and bacteria to human cells~\cite{findlay2014saturation}.
 
@@
 \times 3!} \approx 10^{11}$ triple mutants) make it currently plausible to comprehensively characterize only single mutations to all but the shortest genes.
-Therefore, treating sites independently is not a major limitation for most current datasets; eventually the approach here might be extended to include coupling among mutations.
+Treating sites independently is therefore not a major limitation for most current datasets.
+Eventually the approach here might be extended to include coupling among mutations.
 
-The data for each site $r$ is characterized by the sequencing \emph{depth} (total number of counts); let $N_r^{\rm{pre}}$, $N_r^{\rm{post}}$, $N_r^{s1}$, and $N_r^{s2}$ denote the depth at $r$ for each of the four libraries in Figure \ref{fig:DeepMutationalScan}.
+The data for each site $r$ is characterized by the sequencing \emph{depth} (total number of counts); let $N_r^{\rm{pre}}$, $N_r^{\rm{post}}$, $N_r^{s1}$, and $N_r^{s2}$ denote the depth at $r$ for each of the four libraries in Figure \ref{fig:DeepMutationalScan} (pre-selection, post-selection, selection $s1$, and selection $s2$).
 Typical depths for current experiments are $N \sim 10^6$.
@@
 The values of $n_{r,x}$ for characters $x$ that differ from the wildtype identity $\operatorname{wt}\left(r\right)$ depend on both the depth $N$ and the average per-site mutation rate.
-Typical experiments introduce about one mutation per gene, so $\overline{\mu} \sim 1/L$ where $L$ is the length of the gene.
+Typical experiments introduce about one mutation per gene, so the average mutation rate is $\overline{\mu} \sim 1/L$ where $L$ is the length of the gene.
 Therefore, if a 500-codon gene is sequenced at depth $N \sim 10^6$, we expect $N \overline{\mu} \sim 2000$ counts of non-wildtype codons at each site.
@@
 
-\subsection*{The challenge of statistical inference from finite counts.}
+\subsection*{The goal: inferring site-specific amino-acid preferences.}
 The goal is to estimate the effects of mutations from changes in their counts after selection.
@@
 If we knew these frequencies, we could calculate parameters that reflect the effects of mutations.
-One parameter that characterizes the effect of mutating $r$ from $\operatorname{wt}\left(r\right)$ to $x$ for the experiment in Figure \ref{fig:DeepMutationalScan}A is the \emph{enrichment ratio}
+One parameter that characterizes the effect of mutating $r$ from $\operatorname{wt}\left(r\right)$ to $x$ for the experiment in Figure \ref{fig:DeepMutationalScan}A is the \emph{enrichment ratio}, which is the relative frequency of mutations to $x$ after selection versus before selection:
 \begin{equation}
@@
 \label{eq:preference2}
-f_{r,x} = \frac{\pi_{r,x} \times \mu_{r,x}}{\sum_y \pi_{r,y} \times \mu_{r,y}}.
+f_{r,x} = \frac{\pi_{r,x} \times \mu_{r,x}}{\sum_y \pi_{r,y} \times \mu_{r,y}},
 \end{equation}
+where $y$ is summed over all character identities (all nucleotides, codons, or amino acids).
 The preferences can be intuitively visualized (Figure \ref{fig:DeepMutationalScan}A) and interpreted as the equilibrium frequencies in substitution models for gene evolution~\cite{bloom2014experimentally,bloom2014informed} (after accounting for uneven mutational rates~\cite{yampolsky2005exchangeability,stoltzfus2009climbing}).
 
+\subsection*{The challenge of statistical inference from finite counts.}
 Equations \ref{eq:enrichment} and \ref{eq:preference} are in terms of the true frequencies $\mu_{r,x}$, $f_{r,x}$, etc.
@@
 
-This algorithm a slight modification of that in the \textsl{Methods} of \cite{bloom2014experimentally}; here the algorithm is described anew to explain the implementation in \texttt{dms\_tools}.
+This algorithm is a slight modification of that in the \textsl{Methods} of \cite{bloom2014experimentally}; here the algorithm is described anew to explain the implementation in \texttt{dms\_tools}.
 
 \subsubsection*{Optional controls to quantify error rates.}
-Some sequencing reads that report a mutation may actually reflect an error.
+Some sequencing reads that report a mutation may actually reflect an error introduced during sequencing or PCR rather than an actual mutation that experienced selection.
 Errors can be quantified by sequencing an unmutated gene, so that any counts at $r$ of $x \ne \operatorname{wt}\left(r\right)$ for this control reflect errors.
@@
 For codons, the error rate depends on the number of nucleotides being changed.
-The average error rates  $\overline{\epsilon_m}$ and $\overline{\rho_m}$ for $m$ nucleotide changes are estimated as
+The average error rates $\overline{\epsilon_m}$ and $\overline{\rho_m}$ for codon mutations with $m$ nucleotide changes are estimated as
 \begin{gather}
@@
 
-Our prior assumption is that the mutagenesis introduces all mutant characters at equal frequency (this assumption is only plausible for codons if the mutagenesis is at the codon level as in~\cite{firnberg2012,jain2013rapid,firnberg2014comprehensive,bloom2014experimentally,melnikov2014comprehensive}).
-The per-site mutagenesis rate is estimated as
+Our prior assumption is that the mutagenesis introduces all mutant characters at equal frequency (this assumption is only plausible for codons if the mutagenesis is at the codon level as in~\cite{firnberg2012,jain2013rapid,firnberg2014comprehensive,bloom2014experimentally,melnikov2014comprehensive}; if mutations are made at the nucleotide level such as by error-prone PCR then characters should be defined as nucleotides).
+The average per-site mutagenesis rate is estimated as
 \begin{equation}
@@
 \subsection*{Inferring preferences with \texttt{dms\_tools}.}
+
 \subsubsection*{Application to actual datasets.}
 Figures \ref{fig:MelnikovPrefs} and \ref{fig:ThyagarajanPrefs} illustrate application of \texttt{dms\_tools} to two existing datasets~\cite{melnikov2014comprehensive,thyagarajan2014inherent}.
-As these figures show, the \texttt{dms\_inferprefs} and \texttt{dms\_logoplot} programs infer and visualize the preferences with a few simple commands.
+The programs require as input only simple text files listing the counts of each character identity at each site.
+As the figures show, the \texttt{dms\_inferprefs} and \texttt{dms\_logoplot} programs can process these input files to infer and visualize the preferences with a few simple commands.
 Error controls can be included when available (they are not for Figure \ref{fig:MelnikovPrefs}, but are for Figure \ref{fig:ThyagarajanPrefs}).
@@
 Therefore, the assumption that all mutations are equally tolerated (i.e.
-the preferences for a site are all equal, or the enrichment ratios are all one) is not a plausible null hypothesis for Figure \ref{fig:DeepMutationalScan}A, and so \texttt{dms\_tools} simply infers the preferences given a uniform Dirichlet prior rather than trying to pinpoint some subset of sites with unequal preferences.
+the preferences for a site are all equal, or the enrichment ratios are all one) is not a plausible null hypothesis for Figure \ref{fig:DeepMutationalScan}A.
+For this reason, \texttt{dms\_tools} simply infers the preferences given a uniform Dirichlet prior rather than trying to pinpoint some subset of sites with unequal preferences.
 
-But in Figure \ref{fig:DeepMutationalScan}B, the assumption that most mutations will be similarly selected is a plausible null hypothesis, since only a small subset of mutations will have markedly different effects under two related selections.
+But in Figure \ref{fig:DeepMutationalScan}B, the assumption that most mutations will be similarly selected is a plausible null hypothesis, since we expect to alternative selections to have markedly different effects on only a small subset of mutations (typically, major constraints related to protein folding and stability will be relatively conserved across different selections on the same protein).
 Therefore, \texttt{dms\_tools} uses a different algorithm to infer the differential preferences under the two selections.
@@
 If there is no difference in the effect of $x$ at $r$ between selections $s1$ and $s2$, then $\Delta\pi_{r,x} = 0$.
-If $x$ at $r$ is more preferred by $s2$ than $s1$, then $\Delta\pi_{r,x} > 0$; conversely if $x$ at $r$ is more preferred by $s1$ than $s2$, then $\Delta\pi_{r,x} < 0$.
+If $x$ at $r$ is more preferred by $s2$ than $s1$, then $\Delta\pi_{r,x} > 0$; conversely if $x$ at $r$ is more preferred by $s1$ than $s2$, then $\Delta\pi_{r,x} < 0$ (see Figure \ref{fig:MelnikovSimulatedDiffPrefs}A).
 
@@
 where \texttt{dms\_tools} by default sets all the scalar concentration parameters ($\alpha$'s) to one except $\alpha_{\Delta\pi}$, which is set to two corresponding to a weak expectation that the $\Delta\pi$ values are close to zero.
-The average error rate $\overline{\xi_m}$ for mutations with $m$ of nucleotide changes is
+The average error rate $\overline{\xi_m}$ for mutations with $m$ nucleotide changes is
 \begin{equation}
@@
 The program \texttt{dms\_logoplot} visualizes the posterior mean differential preferences via an extension to \texttt{weblogo}~\cite{crooks2004}.
+In addition, \texttt{dms\_inferdiffprefs} also creates text files that give the posterior probability that $\Delta\pi_{r,x} > 0$ or $< 0$ (in other words, the probability that there is differential selection).
 
@@
 \section*{Acknowledgements}
-Thanks to Alec Heckert for assistance in testing \texttt{dms\_tools}, to Erick Matsen for the excellent suggestion to use \texttt{pystan} for MCMC, and to Nicholas Wu for providing the mutational counts data from \cite{wu2014ns}.
+Thanks to Alec Heckert for assistance in testing \texttt{dms\_tools}, to Erick Matsen for the excellent suggestion to use \texttt{pystan} for MCMC,  to Nicholas Wu for providing the mutational counts data from \cite{wu2014ns}, and to Orr Ashenberg and Hugh Haddox for helpful comments on the manuscript.
 This work was supported by the NIGMS of the NIH under grant R01GM102198.
@@
 \begin{figure}[h!]
-\centerline{\includegraphics[width=6.1in]{Fig_ThyagarajanPrefs/Fig_ThyagarajanPrefs.pdf}}
+\centerline{\includegraphics[width=5.9in]{Fig_ThyagarajanPrefs/Fig_ThyagarajanPrefs.pdf}}
 \cprotect \caption{\label{fig:ThyagarajanPrefs}\csentence{Site-specific preferences from deep mutational scanning of influenza hemagglutinin.} Thyagarajan and Bloom~\cite{thyagarajan2014inherent} performed deep mutational scanning on influenza hemagglutinin, and reported the counts of codon mutations for three biological replicates of the experiment.
@@
 \end{lstlisting}
+Note that unlike in Figure \ref{fig:MelnikovPrefs}, no \texttt{--chartype} option is specified since the \texttt{dms\_inferprefs} default is already \texttt{codon\_to\_aa}.
 }
@@
 \end{lstlisting}
+Note that no \texttt{--chartype} option is specified because the default for \texttt{dms\_inferdiffprefs} is already \texttt{codon\_to\_aa}.
 }

### 2015-01-09 paper/BMC_Bioinformatics_2015/bmc_bioinformatics_article_nofigs.tex
A few more tweaks to paper

--- 2015-01-08 paper/BMC_Bioinformatics_2015/bmc_bioinformatics_article.tex
@@
 
-%\def\includegraphic{}
-%\def\includegraphics{}
+\def\includegraphic{}
+\def\includegraphics{}
 
@@
 \section*{Data and code for figures in this paper.}
-The data and computer code used to generate the figures are in the tagged version of the \texttt{dms\_tools} source code at \url{https://github.com/jbloom/dms_tools/tree/1.0} in \texttt{examples} subdirectory.
+The data and computer code used to generate the figures are in the tagged version of the \texttt{dms\_tools} source code at \url{https://github.com/jbloom/dms_tools} in \texttt{examples} subdirectory.
 The \texttt{LaTex} source for this paper is in the \texttt{paper} subdirectory.
@@
 \begin{figure}[h!]
-\centerline{\includegraphics[clip, trim=0in 0in 5.92in 0in, width=5.5in]{Fig_DeepMutationalScan/Fig_DeepMutationalScan.pdf}}
+%\centerline{\includegraphics[clip, trim=0in 0in 5.92in 0in, width=5.5in]{Fig_DeepMutationalScan/Fig_DeepMutationalScan.pdf}}
 \caption{\label{fig:DeepMutationalScan}{\csentence{A deep mutational scanning experiment.} {\bf (A)} A gene is mutagenized to create a library that contains all single codon mutations.
@@
 \begin{figure}[h!]
-\centerline{\includegraphics[width=6.7in]{Fig_MelnikovPrefs/Fig_MelnikovPrefs.pdf}}
+%\centerline{\includegraphics[width=6.7in]{Fig_MelnikovPrefs/Fig_MelnikovPrefs.pdf}}
 \cprotect \caption{\label{fig:MelnikovPrefs}\csentence{Site-specific preferences from deep mutational scanning of a Tn5 transposon.} Melnikov \textit{et al}~\cite{melnikov2014comprehensive} performed deep mutational scanning on a Tn5 transposon using kanamycin selection, and reported the counts of amino-acid mutations for two biological replicates of the experiment.
@@
 \begin{figure}[h!]
-\centerline{\includegraphics[width=5.9in]{Fig_ThyagarajanPrefs/Fig_ThyagarajanPrefs.pdf}}
+%\centerline{\includegraphics[width=5.9in]{Fig_ThyagarajanPrefs/Fig_ThyagarajanPrefs.pdf}}
 \cprotect \caption{\label{fig:ThyagarajanPrefs}\csentence{Site-specific preferences from deep mutational scanning of influenza hemagglutinin.} Thyagarajan and Bloom~\cite{thyagarajan2014inherent} performed deep mutational scanning on influenza hemagglutinin, and reported the counts of codon mutations for three biological replicates of the experiment.
@@
 \begin{figure}[h!]
-\centerline{\includegraphics[width=6.5in]{../../examples/Melnikov_et_al_Tn5/infer_prefs_on_simulated_data/correlations.pdf}}
+%\centerline{\includegraphics[width=6.5in]{../../examples/Melnikov_et_al_Tn5/infer_prefs_on_simulated_data/correlations.pdf}}
 \cprotect\caption{\label{fig:PrefCorrelations}\csentence{Accuracy of preference inference on simulated data.} Deep mutational scanning counts were simulated using the preferences in Figure \ref{fig:MelnikovPrefs}A and realistic parameters for mutation and error rates.
@@
 \begin{figure}[h!]
-\centerline{\includegraphics[width=6.5in]{Fig_MelnikovDiffPrefs/Fig_MelnikovDiffPrefs.pdf}}
+%\centerline{\includegraphics[width=6.5in]{Fig_MelnikovDiffPrefs/Fig_MelnikovDiffPrefs.pdf}}
 \cprotect\caption{\label{fig:MelnikovSimulatedDiffPrefs}\csentence{Inference of differential preferences on simulated data.} To illustrate and test the inference of differential preferences, the experiment in Figure \ref{fig:DeepMutationalScan}B was simulated at the codon level starting with the post-selection library that yielded the preferences in Figure \ref{fig:MelnikovPrefs}.
@@
 \begin{figure}[h!]
-\centerline{\includegraphics[width=6.5in]{../../examples/Wu_et_al_NS1/diffprefs_logoplot.pdf}}
+%\centerline{\includegraphics[width=6.5in]{../../examples/Wu_et_al_NS1/diffprefs_logoplot.pdf}}
 \cprotect\caption{\label{fig:WuDiffPrefs}\csentence{Differential preferences following selection of influenza NS1 in the presence or absence of interferon.} Wu \textit{et al}~\cite{wu2014ns} generated libraries of influenza viruses carrying nucleotide mutations in the NS segment.

### 2015-03-22 paper/BMC_Bioinformatics_2015/ResponseToReviewers.tex
Added reviewer comments

\documentclass[11pt]{article}
\usepackage{geometry}                % See geometry.pdf to learn the layout options. There are lots.
\geometry{letterpaper}                   % ... or a4paper or a5paper or ... 
\usepackage[parfill]{parskip}    % Activate to begin paragraphs with an empty line rather than an indent
\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{epstopdf}
\usepackage{color}
\DeclareGraphicsRule{.tif}{png}{.png}{`convert #1 `dirname #1`/`basename #1 .tif`.png}

\newcommand{\forceindent}{\leavevmode{\parindent=2em\indent}}

\newcounter{response}
\newenvironment{response}[1][]{\color{blue}\refstepcounter{response}\noindent{{{\bf{Response \arabic{response}.}}}} \normalfont #1}

\renewcommand{\theresponse}{{\bf Response \arabic{response}}}

\usepackage{url}

\title{Response to reviewer comments to the manuscript ``Software for the analysis and visualization of deep mutational scanning data'' submitted to \textit{BMC Bioinformatics}}
\author{Jesse D. Bloom}

\begin{document}

\maketitle

\subsection*{Reviewer \#1}

Bloom describes a software package for analysis of deep mutational scanning using a novel (for this application) Bayesian/MCMC framework. The general framework appears to be sound and addresses a significant need for these new and highly promising functional assays. The manuscript is well-written and explains the major uses cases of the software in sufficient detail.

Minor comments/suggestions:

1. The notion that differences based on larger absolute counts provide more certainty than those based on small absolute counts is obviously true, but only up to a point. Since deep mutational scanning experiments typically involves exponential amplification of a finite number of selected molecules prior to sequencing, there will be a saturation point (or region) beyond which higher absolute counts no longer provide additional information. This effect might not be negligible in experiments that involve typical cultures of $10^6$ to $10^8$ independently transformed or transduced cells. I would encourage the author to note these effects and perhaps add features or provide guidance to software users for how to estimate or detect the saturation point.

2. On pg 2 – the notion of a mutation being ``beneficial'' is generally taken to be relative to the fitness of the wild-type (non-mutated) gene. Eq (1) is such that $\phi_{r,x} > 1$ does not imply that $x$ is beneficial in this sense. Consider the trivial example of a two site gene with two characters. The input library has one member with a neutral mutation at site 1, one member with a deleterious mutation at site 2 and one WT member. Post-selection $\phi$ for the non-WT character at site 1 will be $\frac{1/1}{1/2} > 1$ even though the non-WT character at site 1 is not beneficial. This should be clarified.

3. Given the substantial increase in runtime and computational complexity of the Bayesian/MCMC approach relative to the simple ratio approach (hours vs fraction of a second), it might be helpful to provide a bit more intuition about what is gained. Is it simply a ``regularization'' where high ratios obtained from low counts are pushed back towards the prior? If so, does the framework really provide more than what could be obtained with a much simpler regularization heuristic?

4. It’s not clear whether \textsl{dms\_inferprefs} outputs the probability of $\phi_{r,x} < 1$ (or $\phi_{r,x} \ne 0$). This is likely to be a useful feature.

5. It’s not clear whether the probability values for $\Delta\pi_{r,x} > 0$ or $< 0$ for \textsl{dms\_inferdiffprefs} should be interpreted as nominal or corrected. Providing false discovery rate estimates might be helpful for interpretation (also applies to 4.).

6. The function of \textsl{dms\_merge} isn’t explained in the paper. Is it simply averaging the two replicate profiles? Could additional statistical power be gained by explicitly modeling replicates within the Bayesian framework?

Level of interest: An article of importance in its field

Quality of written English: Acceptable

Statistical review: Yes, and I have assessed the statistics in my report.

\subsection*{Reviewer \#2}

The author describes a Python package, \textsl{dms\_tools}, which is used to infer character preferences using deep mutational scanning data. This software package allows the user to utilize the method described in Bloom 2014 (\textit{Mol Biol Evol}).

I found the paper easy to read and understand, and the algorithm is explained in a clear and logical manner. Simulations show clearly that this method is an improvement over using enrichment ratios to compute site-specific preferences. I have several questions that I would like the authors to comment on, all of which can be considered discretionary revisions:

Question 1:
On page 2, in the second paragraph, a deep mutational scanning study is described. The average mutation rate is $\mu=1/L$, where $L$ is the length of the gene. It is unclear to me why this is so – intuitively one might think that a longer gene may have more mutations, given that mutation is a random process. More generally, could the author provide more details on the range of data one would expect from typical deep mutational scanning experiments?

Question 2:
The author states that for a depth of $10^6$, we expect about 2000 counts of non-wild type codons for the average gene. The author arrived at this number by multiplying the total number of reads by the probability of a mutation. This is reasonable if each read comes from a different DNA fragment, each with probability $\mu$ of mutating. However, if multiple reads all cover the same DNA fragment, would this relationship still hold? The author should comment on whether this situation is likely to occur in practice and, if so, how it affects the method.

Question 3:
Related to question 1: How does the author think that lower-quality data, either by experimental design limitations or by the variance in counts between sites, affect the inference and downstream interoperation of site-specific preferences? How sensitive is the algorithm to the prior for different qualities/quantities of data?

Question 4:
How robust is the method to the prior assumption that mutagenesis introduces all mutations at equal frequency? The author briefly discusses this issue on page 6, and I think it would be informative to address this question with an additional simulation. Does the present method still outperform the naive ratio method even when data are not generated under the model used in the inference?

Level of interest: An article whose findings are important to those with closely related research interests

Quality of written English: Acceptable

Statistical review: No, the manuscript does not need to be seen by a statistician.

\end{document}

### 2015-03-22 paper/BMC_Bioinformatics_2015/ResponseToReviewers.tex
Responded to reviewer 2 comment about uneven mutation rates in simulations

@@
 \newcounter{response}
-\newenvironment{response}[1][]{\color{blue}\refstepcounter{response}\noindent{{{\bf{Response \arabic{response}.}}}} \normalfont #1}
+%\newenvironment{response}[1][]{\color{blue}\refstepcounter{response}\noindent{{{\bf{Response \arabic{response}.}}}} \normalfont #1}
+\newenvironment{response}[1][]{\color{blue}\refstepcounter{response}\noindent \normalfont #1}
 
@@
 \maketitle
+
+\begin{response}
+Thanks to both reviewers for their careful review of the manuscript.
+I appreciate their attention to the manuscript and the software that it describes, and am glad that both reviewers see this software as useful for researchers using the new technique of deep mutational scanning.
+
+The reviewer comments mostly focused on issues of clarification and questions about why and when certain approaches might be preferred.
+I have been able to address all of these comments in the revised manuscript, and I believe that the helpful comments and the changes made in response have substantially improved the manuscript.
+
+Below I provide a point-by-point response to the comments.
+\end{response}
 
@@
 
+\begin{response}
+The reviewer notes that in a real experiment, the per-site per-character mutation ($\mu_{r,x}$) and error rates ($\epsilon_{r,x}$ and $\rho_{r,x}$) will not be completely uniform.
+However, the prior assumption used by \textsl{dms\_tools} is that the these rates are uniform (although this prior is expressed fairly loosely by a Dirichlet with a relatively modest concentration parameter).
+The reviewer is therefore rightly pointing out that if the simulations are performed with uniform mutation (and error) rates, then the inference on the simulated data might work better than it would on real data.
+
+Although the original manuscript text was unclear on this point, in fact the simulations were performed with non-uniform mutation and error rates.
+This was done by drawing the rates from Dirichlet distributions parameterized such that there was substantial variation across sites and characters.
+The unevenness in the simulated rates is controlled by the concentration parameter of the Dirichlet, and the simulations were performed with a concentration parameter that gave an unevenness roughly comparable to that observed in real codon-mutagenesis experiments.
+The code implementing this aspect of the simulations is at \url{????}.
+
+The unevenness in the simulated mutation rates can be grasped in intuitive terms by examining the simulated counts in \url{simulated_errs_depth_2e+06_pre.txt}.
+This simulation is of a library sequenced at depth $2 \times 10^6$ and has an average per-site codon mutation rate of $\overline{\mu} = \frac{1}{264}$, meaning that the average mutation rate to any given one of the 63 mutant codons at any given site is $\frac{1}{264 \times 63} \approx 6 \times 10^{-5}$.
+Therefore, with perfectly uniform mutation rates and no sampling error, each mutant codon would be observed $\frac{2 \times 10^6}{264 \times 63} \approx 120$ times.
+In the simulated data in \url{simulated_errs_depth_2e+06_pre.txt}, the counts for specific mutations range from 0 counts to 1400 counts, with a 10th percentile of 18 counts, a 25th percentile of 43 counts, a 75th percentile of 174 counts, and a 90th percentile of 275 counts.
+Therefore, the unevenness in the simulated data reasonably reflects that of a real experiment, and so these data already represents a fair test for the inferences.
+
+This issue has been clarified in the text by clearly stating that the simulations were done at uneven mutation and error rates.
+\end{response}
+
 Level of interest: An article whose findings are important to those with closely related research interests

### 2015-03-22 paper/BMC_Bioinformatics_2015/bmc_bioinformatics_article.tex
Responded to reviewer 2 comment about uneven mutation rates in simulations

@@
 To test the accuracy of preference-inference by \texttt{dms\_tools}, I simulated deep mutational scanning counts using the preferences in Figure \ref{fig:MelnikovPrefs}, both with and without errors quantified by appropriate controls.
+Importantly, the error and mutation rates for these simulations were \emph{not} uniform across sites and characters, but were simulated to have a level of unevenness comparable to that observed in real experiments.
 I then used \texttt{dms\_tools} to infer preferences from the simulated data, and also made similar inferences using simple ratio estimation (Equation \ref{eq:enrichmentcounts}).
@@
 To test the accuracy of differential preference inference by \texttt{dms\_tools}, I simulated an experiment like that in Figure \ref{fig:DeepMutationalScan}B with the starting counts based on Melnikov \textit{et al}'s actual deep mutational scanning data of a Tn5 transposon~\cite{melnikov2014comprehensive}.
-As show by Figure \ref{fig:MelnikovSimulatedDiffPrefs}, \texttt{dms\_inferdiffprefs} accurately infers the differential preferences at typical experimental depths.
+As shown by Figure \ref{fig:MelnikovSimulatedDiffPrefs}, \texttt{dms\_inferdiffprefs} accurately infers the differential preferences at typical experimental depths.
 The results are easily visualized with \texttt{dms\_logoplot}.
@@
 \centerline{\includegraphics[width=6.5in]{../../examples/Melnikov_et_al_Tn5/infer_prefs_on_simulated_data/correlations.pdf}}
-\cprotect\caption{\label{fig:PrefCorrelations}\csentence{Accuracy of preference inference on simulated data.} Deep mutational scanning counts were simulated using the preferences in Figure \ref{fig:MelnikovPrefs}A and realistic parameters for mutation and error rates.
+\cprotect\caption{\label{fig:PrefCorrelations}\csentence{Accuracy of preference inference on simulated data.} Deep mutational scanning counts were simulated using the preferences in Figure \ref{fig:MelnikovPrefs}A and realistic mutation and error rates that were uneven across sites and characters as in actual experiments.
 The simulations were done {\bf (A)} without or {\bf (B)} with sequencing errors quantified by control libraries.

### 2015-03-22 paper/BMC_Bioinformatics_2015/ResponseToReviewers.tex
Responded to reviewer 2 about average mutation rate

@@
 The average mutation rate is $\mu=1/L$, where $L$ is the length of the gene.
-It is unclear to me why this is so – intuitively one might think that a longer gene may have more mutations, given that mutation is a random process.
+It is unclear to me why this is so -- intuitively one might think that a longer gene may have more mutations, given that mutation is a random process.
 More generally, could the author provide more details on the range of data one would expect from typical deep mutational scanning experiments?
+
+\begin{response}
+In principle, a deep mutational scanning experiment could introduce any number of mutations per gene.
+But typically, experimentalists tend to introduce an average of $\approx$1 mutation per gene when they create the codon-mutant libraries.
+The reason is that having genes with $\gg$1 mutations makes it hard to attribute the selection on any given gene variant to any specific mutation.
+For this reason, published deep mutational scanning experiments have generally targeted $\overline{\mu} \sim 1/L$.
+This is an important distinction between deep mutational scanning and naturally occurring mutations -- naturally occurring mutations will be more common in longer genes, since typically the per-site mutation rate is approximately uniform and longer genes have more sites.
+The revised text now clarifies this point.
+\end{response}
 

### 2015-03-22 paper/BMC_Bioinformatics_2015/bmc_bioinformatics_article.tex
Responded to reviewer 2 about average mutation rate

@@
 Denote the counts of character $x$ (characters might be nucleotides, amino acids, or codons) at $r$ as $n_{r,x}^{\rm{pre}}$, $n_{r,x}^{\rm{post}}$, $n_{r,x}^{s1}$, and $n_{r,x}^{s2}$.
-The values of $n_{r,x}$ for characters $x$ that differ from the wildtype identity $\operatorname{wt}\left(r\right)$ depend on both the depth $N$ and the average per-site mutation rate.
-Typical experiments introduce about one mutation per gene, so the average mutation rate is $\overline{\mu} \sim 1/L$ where $L$ is the length of the gene.
+The values of $n_{r,x}$ for characters $x$ that differ from the wildtype identity $\operatorname{wt}\left(r\right)$ depend on both the depth $N$ and the average per-site mutation rate $\overline{\mu}$.
+Since the mutations are intentionally introduced into the mutant library by the experimentalist, in principle $\overline{\mu}$ could have any value.
+But typically, deep mutational scanning experiments aim to introduce about one mutation per gene to avoid filling the mutant library with highly mutated genes -- so the average mutation rate is usually $\overline{\mu} \sim 1/L$ where $L$ is the length of the gene.
 Therefore, if a 500-codon gene is sequenced at depth $N \sim 10^6$, we expect $N \overline{\mu} \sim 2000$ counts of non-wildtype codons at each site.

### 2015-03-23 paper/BMC_Bioinformatics_2015/ResponseToReviewers.tex
Responded to reviewer 1 comment 1, reviewer 2 comment 2

@@
 I would encourage the author to note these effects and perhaps add features or provide guidance to software users for how to estimate or detect the saturation point.
+
+\begin{response}
+This is an excellent point, and is completely true!
+The text has now been revised to emphasize that increasing deep-sequencing counts stop being meaningful once the typical number of counts per site becomes on the order of the number of unique molecules that were originally in the experiment.
+For most shotgun sequencing experiments, the typical counts per site ($\sim 10^5$ to $10^6$) still tend to be less than the typical number of unique molecules ($\sim 10^6$ to $10^8$), but this will not always be true particularly as sequencing depths increase.
+
+The revised text now emphasizes this point.
+It emphasizes that the experiments should be done in a way to ensure that the sequencing depth is not saturating the number of unique molecules (i.e.
+by quantifying the number of molecules carefully).
+It also suggests that one way that this issue can be examined during \textit{post hoc} computational analyses is by sub-sampling the deep sequencing data (i.e.
+randomly selecting only a portion of the sequencing reads) when generated the mutation counts files used as input to inference programs in \textsl{dms\_tools}, and then seeing if this reduced depth alters the results.
+When the results start to become insensitive to sub-sampling the data, this provides an indication that the unique molecules are starting to be oversampled.
+\end{response}
 
@@
 
+\begin{response}
+This is an excellent comment, and is closely related to point (1) made by the first reviewer.
+The calculation to of the counts to which the reviewer refers, and indeed much of the statistical framework described in the manuscript, is made under the assumption that the number of unique DNA fragments in the library substantially exceeds the number of sequencing reads, such that the deep-sequencing reads can be viewed as subsampling a very large number of DNA fragments.
+This issue has been clarified in the text as described in the response to point (1) made by the first reviewer.
+\end{response}
+
 Question 3:

### 2015-03-23 paper/BMC_Bioinformatics_2015/bmc_bioinformatics_article.tex
Responded to reviewer 1 comment 1, reviewer 2 comment 2

@@
 Therefore, even at depths $N \sim 10^6$, the actual counts of most mutations will be quite modest.
+
+The rest of this paper assumes that the sequencing depth is less than the number of unique molecules in the mutant library, such that the deep sequencing randomly subsamples the set of molecules.
+If this assumption is false (i.e.
+if the number of unique molecules is substantially less than the sequencing depth), then the accuracy of inferences about mutational effects will be fundamentally limited by this aspect of the experimental design.
+Properly done experiments should quantify the number of unique molecules in the library so that it is obvious whether this assumption holds.
+In the absence of such information, the analysis can be repeated using only a random fraction of the deep sequencing data to assess whether inferences are limited by sequencing depth or they underlying molecular diversity in the mutant library.
 

### 2015-03-23 paper/BMC_Bioinformatics_2015/ResponseToReviewers.tex
Responded to reviewer 1 comment 2

--- 2015-03-22 paper/BMC_Bioinformatics_2015/ResponseToReviewers.tex
@@
 
+\begin{response}
+This is an excellent point.
+The text has been revised to emphasize that increasing deep-sequencing counts stop being meaningful once the typical number of counts per site becomes on the order of the number of unique molecules that were originally in the experiment.
+For most shotgun sequencing experiments, the typical counts per site ($\sim 10^5$ to $10^6$) still tend to be less than the typical number of unique molecules ($\sim 10^6$ to $10^8$), but this may not always be true.
+The revised text suggests that the experiments should be done in a way that ensures that the sequencing depth is not saturating the number of unique molecules (i.e.
+by quantifying the number of molecules carefully).
+It suggests that one way that this issue can be examined during \textit{post hoc} computational analyses is by randomly subsampling the deep sequencing data (when generating the mutation counts files used as input to inference programs in \textsl{dms\_tools}), and then seeing if this reduced depth alters the results.
+\end{response}
+
 2.
@@
 This should be clarified.
+
+\begin{response}
+This is a \emph{hugely} important point!
+The first paragraph of the subsection in the \textsl{Background} entitled \textsl{The goal: inferring site-specific amino-acid preferences} now discusses this issue in detail.
+Essentially, assuming the length $L$ of the gene is $\gg 1$ (which is the case for any reasonable gene), then if the mutant library contains a Poisson distribution of the number of mutations per gene (as is often the case), the problem noted by the reviewer will not arise in practice.
+The reason is that for such genes, knowing that there is a mutation at site $r$ does not appreciably change the marginal distribution of the number of mutations at other sites (it only changes it by a factor of $\frac{L - 1}{L}$, which in general will be very small).
+On the other hand, if the mutant library is created such that it contains genes with exactly one mutation per gene (which is now sometimes done), then the problem noted by the reviewer will arise even for realistic genes with $L \gg 1$.
+The subsection \textsl{The goal: inferring site-specific amino-acid preferences} discusses this issue, and how the problem can be corrected by appropriately defining the frequencies for the wildtype character.
+\end{response}
 
@@
 
+\begin{response}
+This is an excellent comment, and is closely related to point (1) made by the first reviewer.
+The calculation of the counts to which the reviewer refers, and indeed much of the statistical framework described in the manuscript, is made under the assumption that the number of unique DNA fragments in the library substantially exceeds the number of sequencing reads, such that the deep-sequencing reads can be viewed as subsampling a very large number of DNA fragments.
+This issue has been clarified in the text as described in the response to point (1) made by the first reviewer.
+\end{response}
+
 Question 3:

### 2015-03-23 paper/BMC_Bioinformatics_2015/bmc_bioinformatics_article.tex
Responded to reviewer 1 comment 2

@@
 Deep sequencing of the variants pre- and post-selection provides information about the functional impacts of mutations.
-Since the original description of deep mutational scanning by Fowler \textit{et al}~\cite{fowler2010high}, the technique has been applied to a wide range of genes~\cite{traxlmayr2012construction,mclaughlin2012spatial,starita2013activity,melamed2013deep,roscoe2013analyses,firnberg2014comprehensive,bloom2014experimentally,melnikov2014comprehensive,thyagarajan2014inherent,wu2014high,wu2014ns,olson2014comprehensive}, both to measure mutational tolerance given a single selection pressure as in Figure \ref{fig:DeepMutationalScan}A, or to identify mutations that have different effects under alternative selections as in Figure \ref{fig:DeepMutationalScan}B.
-New techniques to create comprehensive codon-mutant libraries of genes make it possible to profile all single amino-acid mutations~\cite{firnberg2012,jain2013rapid,firnberg2014comprehensive,bloom2014experimentally,melnikov2014comprehensive}, while new techniques for targeted mutagenesis of mammalian genomes enable deep mutational scanning to be applied across the biological spectrum from viruses and bacteria to human cells~\cite{findlay2014saturation}.
+Since the original description of deep mutational scanning by Fowler \textit{et al}~\cite{fowler2010high}, the technique has been applied to a wide range of genes~\cite{traxlmayr2012construction,mclaughlin2012spatial,starita2013activity,melamed2013deep,roscoe2013analyses,firnberg2014comprehensive,bloom2014experimentally,melnikov2014comprehensive,thyagarajan2014inherent,wu2014high,wu2014ns,olson2014comprehensive,kitzman2015massively}, both to measure mutational tolerance given a single selection pressure as in Figure \ref{fig:DeepMutationalScan}A, or to identify mutations that have different effects under alternative selections as in Figure \ref{fig:DeepMutationalScan}B.
+New techniques to create comprehensive codon-mutant libraries of genes make it possible to profile all single amino-acid mutations~\cite{firnberg2012,jain2013rapid,firnberg2014comprehensive,bloom2014experimentally,melnikov2014comprehensive,kitzman2015massively}, while new techniques for targeted mutagenesis of mammalian genomes enable deep mutational scanning to be applied across the biological spectrum from viruses and bacteria to human cells~\cite{findlay2014saturation}.
 
@@
 The goal is to estimate the effects of mutations from changes in their counts after selection.
-Let $\mu_{r,x}$, $f_{r,x}$, $f_{r,x}^{s1}$, and $f_{r,x}^{s2}$ denote the \emph{true} frequencies of $x$ at $r$ that would be observed for the four libraries in Figure \ref{fig:DeepMutationalScan} if we sampled at infinite depth in both the actual experiment and the sequencing.
-If we knew these frequencies, we could calculate parameters that reflect the effects of mutations.
+Let $\mu_{r,x}$, $f_{r,x}$, $f_{r,x}^{s1}$, and $f_{r,x}^{s2}$ denote the \emph{true} frequencies at site $r$ of all mutant characters $x \ne \operatorname{wt}\left(r\right)$ that would be observed for the four libraries in Figure \ref{fig:DeepMutationalScan} if we sampled at infinite depth in both the actual experiment and the sequencing.
+The definition of these frequencies for the wildtype character $\operatorname{wt}\left(r\right)$ depends on how the mutant library is constructed.
+If the mutant library is constructed so that there is a Poisson distribution of the number of mutations per gene (as is the case for error-prone PCR or the codon-mutagenesis in \cite{bloom2014experimentally,thyagarajan2014inherent}), then $\mu_{r,\operatorname{wt}\left(r\right)}$, $f_{r,\operatorname{wt}\left(r\right)}$, $f_{r,\operatorname{wt}\left(r\right)}^{s1}$, and $f_{r,\operatorname{wt}\left(r\right)}^{s2}$ are defined as for all other characters $x$, and denote the frequencies of $\operatorname{wt}\left(r\right)$ at site $r$ that would be observed if sampling at infinite depth.
+The reason we can make this definition for libraries containing genes with Poisson-distributed numbers of mutations is that for any reasonable-length gene ($L \gg 1$), the marginal distribution of the number of mutations in a gene is virtually unchanged by the knowledge that there is a mutation at site $r$.
+On the other hand, if the mutant library is constructed so that there is exactly zero or one mutation per gene (as in \cite{firnberg2014comprehensive,melnikov2014comprehensive,kitzman2015massively}), then the marginal distribution of the total number of mutations in a gene is changed by the knowledge that there is a mutation at $r$.
+In this case, the wildtype-character frequencies $\mu_{r,\operatorname{wt}\left(r\right)}$, $f_{r,\operatorname{wt}\left(r\right)}$, $f_{r,\operatorname{wt}\left(r\right)}^{s1}$, and $f_{r,\operatorname{wt}\left(r\right)}^{s2}$ are correctly defined as the frequency of unmutated genes in the library, and the counts $n_{r,\operatorname{wt}\left(r\right)}^{\rm{pre}}$, etc.
+are defined as the number of reads at $r$ attributable to unmutated genes.
+In this case, measurement of these counts requires sequencing with linkage as in \cite{hiatt2010parallel,wu2014hiv,kitzman2015massively}.
+The proper analysis of libraries containing only unmutated and singly mutated clones sequenced without linkage is beyond the scope of this paper.
+
+If we knew the frequencies $\mu_{r,x}$, $f_{r,x}$, $f_{r,x}^{s1}$, and $f_{r,x}^{s2}$, we could calculate parameters that reflect the effects of mutations.
 One parameter that characterizes the effect of mutating $r$ from $\operatorname{wt}\left(r\right)$ to $x$ for the experiment in Figure \ref{fig:DeepMutationalScan}A is the \emph{enrichment ratio}, which is the relative frequency of mutations to $x$ after selection versus before selection:
@@
 
-Our prior assumption is that the mutagenesis introduces all mutant characters at equal frequency (this assumption is only plausible for codons if the mutagenesis is at the codon level as in~\cite{firnberg2012,jain2013rapid,firnberg2014comprehensive,bloom2014experimentally,melnikov2014comprehensive}; if mutations are made at the nucleotide level such as by error-prone PCR then characters should be defined as nucleotides).
+Our prior assumption is that the mutagenesis introduces all mutant characters at equal frequency (this assumption is only plausible for codons if the mutagenesis is at the codon level as in~\cite{firnberg2012,jain2013rapid,firnberg2014comprehensive,bloom2014experimentally,melnikov2014comprehensive,kitzman2015massively}; if mutations are made at the nucleotide level such as by error-prone PCR then characters should be defined as nucleotides).
 The average per-site mutagenesis rate is estimated as

### 2015-03-24 paper/BMC_Bioinformatics_2015/ResponseToReviewers.tex
Responded to reviewer 1 comment 6

@@
 
-4. It’s not clear whether \textsl{dms\_inferprefs} outputs the probability of $\phi_{r,x} < 1$ (or $\phi_{r,x} \ne 0$). This is likely to be a useful feature.
+4. It’s not clear whether \texttt{dms\_inferprefs} outputs the probability of $\phi_{r,x} < 1$ (or $\phi_{r,x} \ne 0$). This is likely to be a useful feature.
 
 5.
-It’s not clear whether the probability values for $\Delta\pi_{r,x} > 0$ or $< 0$ for \textsl{dms\_inferdiffprefs} should be interpreted as nominal or corrected.
+It's not clear whether the probability values for $\Delta\pi_{r,x} > 0$ or $< 0$ for \texttt{dms\_inferdiffprefs} should be interpreted as nominal or corrected.
 Providing false discovery rate estimates might be helpful for interpretation (also applies to 4.).
 
+\begin{response}
+The posterior probabilities $\Pr\left(\Delta\pi_{r,x} > 0 \mid \mbox{data}\right)$ reported by \texttt{dms\_inferdiffprefs} are \emph{not} corrected.
+However, note that these are Bayesian posterior probabilities, not classical $P$-values.
+By default, \texttt{dms\_inferdiffprefs} places a prior over the $\Delta\pi_{r,x}$ values that biases them towards zero (see Equation 31 of the manuscript).
+Therefore, it is not clear that these posterior probabilities need to be corrected -- there are arguments for why Bayesian posterior probabilities inferred with appropriate priors do not need to be corrected (see \url{http://www.stat.columbia.edu/~gelman/research/published/multiple2f.pdf}).
+Rather than discuss these complex and incompletely resolved statistical questions, we have simply added the following sentence:
+\begin{quote}
+In addition, \texttt{dms\_inferdiffprefs} also creates text files that give the posterior probability that $\Delta\pi_{r,x} > 0$ or $< 0$ (these posterior probabilities are \emph{not} corrected to account for the fact that multiple sites are typically being examined, although be default the inferences are made using the regularizing prior in Equation 31).
+\end{quote}
+\end{response}
+
 6.
-The function of \textsl{dms\_merge} isn’t explained in the paper.
+The function of \textsl{dms\_merge} isn't explained in the paper.
 Is it simply averaging the two replicate profiles?
@@
 
-The author describes a Python package, \textsl{dms\_tools}, which is used to infer character preferences using deep mutational scanning data.
+The author describes a Python package, \texttt{dms\_tools}, which is used to infer character preferences using deep mutational scanning data.
 This software package allows the user to utilize the method described in Bloom 2014 (\textit{Mol Biol Evol}).

### 2015-03-24 paper/BMC_Bioinformatics_2015/bmc_bioinformatics_article.tex
Responded to reviewer 1 comment 6

@@
 The program \texttt{dms\_logoplot} visualizes the posterior mean differential preferences via an extension to \texttt{weblogo}~\cite{crooks2004}.
-In addition, \texttt{dms\_inferdiffprefs} also creates text files that give the posterior probability that $\Delta\pi_{r,x} > 0$ or $< 0$ (in other words, the probability that there is differential selection).
+In addition, \texttt{dms\_inferdiffprefs} also creates text files that give the posterior probability that $\Delta\pi_{r,x} > 0$ or $< 0$ (these posterior probabilities are \emph{not} corrected to account for the fact that multiple sites are typically being examined, although be default the inferences are made using the regularizing prior in Equation \ref{eq:pr_deltapi}).
 

### 2015-03-24 paper/BMC_Bioinformatics_2015/ResponseToReviewers.tex
Responded to reviewer 1 comment 3

@@
 
+\begin{response}
+The reviewer asks the eminently reasonable question, ``Is the computational complexity of the Bayesian inference worth it?'' The revised manuscript includes a new subsection entitled \textsl{Is the Bayesian inference worthwhile?} directly after the description of the preference inference on simulated data.
+The honest answer to this question is equivocal -- the Bayesian inference appears to always perform as well or better than simple ratio calculations, but the improvements are non-negligible only in some situations.
+Hopefully the discussion in this new subsection provides a fair discussion of this topic.
+
+Note that even when ratio estimation (which is also implemented in \texttt{dms\_tools} via the \texttt{--ratio\_estimation} option) is just as good, the visualization capabilities of \texttt{dms\_tools} will still be useful.
+\end{response}
+
 4. It’s not clear whether \texttt{dms\_inferprefs} outputs the probability of $\phi_{r,x} < 1$ (or $\phi_{r,x} \ne 0$). This is likely to be a useful feature.
@@
 6.
-The function of \textsl{dms\_merge} isn't explained in the paper.
+The function of \texttt{dms\_merge} isn't explained in the paper.
 Is it simply averaging the two replicate profiles?

### 2015-03-24 paper/BMC_Bioinformatics_2015/bmc_bioinformatics_article.tex
Responded to reviewer 1 comment 3

@@
 
+\subsubsection*{Is the Bayesian inference worthwhile?}
+The foregoing sections explain why the Bayesian inference of preferences implemented in \texttt{dms\_tools} is conceptually preferable to estimating mutational effects via direct calculation of enrichment ratios using Equation \ref{eq:enrichmentcounts}.
+However, do the practical benefits of this Bayesian inference justify its increased complexity?
+The simulations in the previous section show that the Bayesian inference is more accurate, but in the absence of background errors (Figure \ref{fig:PrefCorrelations}A) the magnitude of the improvement becomes negligible once the mutations counts per site $N \overline{\mu}$ start to exceed $\sim10^3$.
+When there is a need to correct for background errors (Figure \ref{fig:PrefCorrelations}B), meaningful benefits of the Bayesian inference over enrichment ratios extend to somewhat higher sequencing depths.
+Overall, it appears that Bayesian inference will always perform as well or better than enrichment ratios.
+But the tangible benefit becomes negligible at high sequencing depth.
+In that case, the user will have to decide if the increased computational runtime and complexity of the Bayesian inference is worth a small improvement in accuracy.
+Simpler enrichment-ratio calculations can be performed using the \texttt{--ratio\_estimation} option of \texttt{dms\_inferprefs} or using an alternative program such as \texttt{Enrich}~\cite{fowler2011enrich}.
+When applying enrichment-ratio calculations to data where some mutations have low counts, it is important to include pseudocounts (denoted by $\mathcal{P}$ in Equation \ref{eq:enrichmentcounts}) as a form of regularization to avoid estimating excessively high or low ratios at sites with limited counts.
+
 \subsection*{Algorithm to infer differential preferences.}

### 2015-03-24 paper/BMC_Bioinformatics_2015/ResponseToReviewers.tex
Responded to reviewer 1 comment 4

@@
 
-4. It’s not clear whether \texttt{dms\_inferprefs} outputs the probability of $\phi_{r,x} < 1$ (or $\phi_{r,x} \ne 0$). This is likely to be a useful feature.
+4. It’s not clear whether \texttt{dms\_inferprefs} outputs the probability of $\phi_{r,x} < 1$ (or $\phi_{r,x} \ne 1$). This is likely to be a useful feature.
+
+\begin{response}
+The \texttt{dms\_inferprefs} program does not output the posterior probability of $\phi_{r,x} \ne 1$ (or $< 1$), but it does output the 95\% credible interval for each $\pi_{r,x}$ value.
+The reason for not outputting the posterior probabilities of $\phi_{r,x} \ne 1$ is that the hypothesis that all amino acids have equal preferences at a given site does not represent a realistic view of proteins, as it is doubtful that there is any site in any protein where all 20 amino acids are equally preferred.
+Therefore, rather than providing a posterior probability that such an unrealistic hypothesis is true, it is more informative to simply output the credible interval of the preference for each amino acid at each site.
+This situation differs for the differential preferences discussed in the reviewer's next point -- for the differential preferences, it \emph{is} a plausible hypothesis that most preferences will be equal under two selection pressures, so computing posterior probabilities about this hypothesis is potentially useful.
+\end{response}
 
@@
 The posterior probabilities $\Pr\left(\Delta\pi_{r,x} > 0 \mid \mbox{data}\right)$ reported by \texttt{dms\_inferdiffprefs} are \emph{not} corrected.
-However, note that these are Bayesian posterior probabilities, not classical $P$-values.
+However, note that these are Bayesian posterior probabilities, \emph{not} classical $P$-values.
 By default, \texttt{dms\_inferdiffprefs} places a prior over the $\Delta\pi_{r,x}$ values that biases them towards zero (see Equation 31 of the manuscript).

### 2015-03-24 paper/BMC_Bioinformatics_2015/ResponseToReviewers.tex
Responded to reviewer 1 point 6

@@
 
+\begin{response}
+The revised manuscript now briefly explains \texttt{dms\_merge} and \texttt{dms\_correlate}.
+To jointly analyze replicates of the same experiment within the Bayesian framework, the counts from the replicates can simply be summed before inputting them to \texttt{dms\_inferprefs}.
+However, it may sometimes be desirable to analyze replicates individually with \texttt{dms\_inferprefs} and then combine and compare them using \texttt{dms\_merge} and \texttt{dms\_correlate}.
+The reason is that some of the replicate-to-replicate variability may arise from experimental factors distinct from statistical noise -- such factors are not easily incorporated into the Bayesian framework, since they are due to experimental rather than statistical sources.
+\end{response}
+
 Level of interest: An article of importance in its field

### 2015-03-24 paper/BMC_Bioinformatics_2015/bmc_bioinformatics_article.tex
Responded to reviewer 1 point 6

@@
 The program \texttt{dms\_logoplot} in the \texttt{dms\_tools} package visualizes the posterior mean preferences via an extension to \texttt{weblogo}~\cite{crooks2004}.
+The program \texttt{dms\_merge} can be used to average preferences inferred from different experimental replicates that have individually been analyzed by \texttt{dms\_inferprefs}, and the program \texttt{dms\_correlate} can be used to compute the correlations among inferences from different replicates.
 

### 2015-03-24 paper/BMC_Bioinformatics_2015/ResponseToReviewers.tex
Responded to remaining comments from reviewer 2

@@
 
+\usepackage{hyperref}
 \usepackage{url}
+\usepackage{breakurl}
 
@@
 Question 3:
-Related to question 1: How does the author think that lower-quality data, either by experimental design limitations or by the variance in counts between sites, affect the inference and downstream interoperation of site-specific preferences?
+Related to question 1: How does the author think that lower-quality data, either by experimental design limitations or by the variance in counts between sites, affect the inference and downstream interpretation of site-specific preferences?
 How sensitive is the algorithm to the prior for different qualities/quantities of data?
+
+\begin{response}
+Lower-quality data does slightly degrade the performance of the inference.
+For instance, comparison of Figure 4A and Figure 4B shows that the presence of background errors in the counts requires the use of greater sequencing depth to make accurate inferences.
+But an advantage of \texttt{dms\_tools} is that it can correct for these errors (given appropriate controls) -- and so in the presence of errors its improvement over the simpler approach of ratio estimation is greater.
+Obviously if there are errors that cannot be quantified by proper controls, \texttt{dms\_tools} will be misled as will any inference approach.
+
+The algorithm is not particularly sensitive to the priors used.
+In developing the software, a variety of concentration parameters for the Dirichlet priors were evaluated.
+The conclusion was that uniform Dirichlets (i.e.
+concentration parameters of one) that treat all possible preference vectors as equally probable \textit{a priori} gave the most robust performance over a wide range of parameters.
+These priors are therefore the ones implemented by default in \texttt{dms\_tools}, and are the ones that were used in the manuscript.
+The programs do enable the user to try other concentration parameters for the priors by following the instructions detailed in the manual at \url{http://jbloom.github.io/dms_tools/}.
+
+As noted in the response to the reviewers next question, the simulations were performed to match realistic properties of a real experiment.
+\end{response}
 
@@
 The unevenness in the simulated rates is controlled by the concentration parameter of the Dirichlet, and the simulations were performed with a concentration parameter that gave an unevenness roughly comparable to that observed in real codon-mutagenesis experiments.
-The code implementing this aspect of the simulations is at \url{????}.
+The code implementing this aspect of the simulations is at \url{https://github.com/jbloom/dms_tools/blob/master/src/simulate.py#L169}.
 
-The unevenness in the simulated mutation rates can be grasped in intuitive terms by examining the simulated counts in \url{simulated_errs_depth_2e+06_pre.txt}.
+The unevenness in the simulated mutation rates can be grasped in intuitive terms by examining the simulated counts in \url{https://github.com/jbloom/dms_tools/blob/master/examples/Melnikov_et_al_Tn5/infer_prefs_on_simulated_data/simulated_data/simulated_depth_2e%2B06_pre.txt}.
 This simulation is of a library sequenced at depth $2 \times 10^6$ and has an average per-site codon mutation rate of $\overline{\mu} = \frac{1}{264}$, meaning that the average mutation rate to any given one of the 63 mutant codons at any given site is $\frac{1}{264 \times 63} \approx 6 \times 10^{-5}$.
 Therefore, with perfectly uniform mutation rates and no sampling error, each mutant codon would be observed $\frac{2 \times 10^6}{264 \times 63} \approx 120$ times.
-In the simulated data in \url{simulated_errs_depth_2e+06_pre.txt}, the counts for specific mutations range from 0 counts to 1400 counts, with a 10th percentile of 18 counts, a 25th percentile of 43 counts, a 75th percentile of 174 counts, and a 90th percentile of 275 counts.
+In the simulated data, the counts for specific mutations range from 0 counts to 1400 counts, with a 10th percentile of 18 counts, a 25th percentile of 43 counts, a 75th percentile of 174 counts, and a 90th percentile of 275 counts.
 Therefore, the unevenness in the simulated data reasonably reflects that of a real experiment, and so these data already represents a fair test for the inferences.

### 2015-03-24 paper/BMC_Bioinformatics_2015/ResponseToReviewers.tex
Final check of reviewer responses

@@
 Thanks to both reviewers for their careful review of the manuscript.
-I appreciate their attention to the manuscript and the software that it describes, and am glad that both reviewers see this software as useful for researchers using the new technique of deep mutational scanning.
+I appreciate their attention, and am glad that both reviewers see this software as useful for researchers using the new technique of deep mutational scanning.
 
-The reviewer comments mostly focused on issues of clarification and questions about why and when certain approaches might be preferred.
-I have been able to address all of these comments in the revised manuscript, and I believe that the helpful comments and the changes made in response have substantially improved the manuscript.
+The reviewers mostly focused on issues of clarification and questions about why and when certain approaches might be preferred.
+I have addressed all of their comments in the revised manuscript, and the helpful comments have substantially improved the manuscript.
 
-Below I provide a point-by-point response to the comments.
+Below I provide a point-by-point response that details the changes to the revised manuscript.
 \end{response}
@@
 \begin{response}
-This is a \emph{hugely} important point!
-The first paragraph of the subsection in the \textsl{Background} entitled \textsl{The goal: inferring site-specific amino-acid preferences} now discusses this issue in detail.
+This is a \emph{hugely} important point that was not adequately covered in the original manuscript!
+The original manuscript failed to describe caveats related to cases such as the one raised by the reviewer.
+
+In the revised manuscript, the first paragraph of the subsection in \textsl{Background} entitled \textsl{The goal: inferring site-specific amino-acid preferences} now discusses this issue in detail.
 Essentially, assuming the length $L$ of the gene is $\gg 1$ (which is the case for any reasonable gene), then if the mutant library contains a Poisson distribution of the number of mutations per gene (as is often the case), the problem noted by the reviewer will not arise in practice.
@@
 The reviewer asks the eminently reasonable question, ``Is the computational complexity of the Bayesian inference worth it?'' The revised manuscript includes a new subsection entitled \textsl{Is the Bayesian inference worthwhile?} directly after the description of the preference inference on simulated data.
-The honest answer to this question is equivocal -- the Bayesian inference appears to always perform as well or better than simple ratio calculations, but the improvements are non-negligible only in some situations.
-Hopefully the discussion in this new subsection provides a fair discussion of this topic.
+The honest answer to this question is equivocal -- the Bayesian inference always performs as well or better than simple ratio calculations that are regularized by the addition of pseudocounts, but the improvements are non-negligible only in some situations.
+The new subsection provides a fair discussion of this topic.
 
-Note that even when ratio estimation (which is also implemented in \texttt{dms\_tools} via the \texttt{--ratio\_estimation} option) is just as good, the visualization capabilities of \texttt{dms\_tools} will still be useful.
+Note that even when ratio estimation (which is also implemented in \texttt{dms\_tools} via the \texttt{--ratio\_estimation} option) performs just as well as the Bayesian inference, the visualization capabilities of \texttt{dms\_tools} will still be useful.
 \end{response}
@@
 The reason for not outputting the posterior probabilities of $\phi_{r,x} \ne 1$ is that the hypothesis that all amino acids have equal preferences at a given site does not represent a realistic view of proteins, as it is doubtful that there is any site in any protein where all 20 amino acids are equally preferred.
-Therefore, rather than providing a posterior probability that such an unrealistic hypothesis is true, it is more informative to simply output the credible interval of the preference for each amino acid at each site.
+Therefore, rather than providing a posterior probability that such an unrealistic hypothesis is true (or false), it is more informative to simply output the credible interval of the preference for each amino acid at each site.
 This situation differs for the differential preferences discussed in the reviewer's next point -- for the differential preferences, it \emph{is} a plausible hypothesis that most preferences will be equal under two selection pressures, so computing posterior probabilities about this hypothesis is potentially useful.
@@
 Therefore, it is not clear that these posterior probabilities need to be corrected -- there are arguments for why Bayesian posterior probabilities inferred with appropriate priors do not need to be corrected (see \url{http://www.stat.columbia.edu/~gelman/research/published/multiple2f.pdf}).
-Rather than discuss these complex and incompletely resolved statistical questions, we have simply added the following sentence:
+Rather than discuss these complex and incompletely resolved statistical questions, we have simply added the following sentences:
 \begin{quote}
-In addition, \texttt{dms\_inferdiffprefs} also creates text files that give the posterior probability that $\Delta\pi_{r,x} > 0$ or $< 0$ (these posterior probabilities are \emph{not} corrected to account for the fact that multiple sites are typically being examined, although be default the inferences are made using the regularizing prior in Equation 31).
+In addition, \texttt{dms\_inferdiffprefs} creates text files that give the posterior probability that $\Delta\pi_{r,x} > 0$ or $< 0$.
+These posterior probabilities are \emph{not} corrected to account for the fact that multiple sites are typically being examined, although be default the inferences are made using the regularizing prior in Equation 31.
 \end{quote}
@@
 The reviewer notes that in a real experiment, the per-site per-character mutation ($\mu_{r,x}$) and error rates ($\epsilon_{r,x}$ and $\rho_{r,x}$) will not be completely uniform.
-However, the prior assumption used by \textsl{dms\_tools} is that the these rates are uniform (although this prior is expressed fairly loosely by a Dirichlet with a relatively modest concentration parameter).
+However, the prior assumption used by \texttt{dms\_tools} is that each rate is equal to the mean (although this prior is expressed only loosely by a Dirichlet with a modest concentration parameter).
 The reviewer is therefore rightly pointing out that if the simulations are performed with uniform mutation (and error) rates, then the inference on the simulated data might work better than it would on real data.
@@
 In the simulated data, the counts for specific mutations range from 0 counts to 1400 counts, with a 10th percentile of 18 counts, a 25th percentile of 43 counts, a 75th percentile of 174 counts, and a 90th percentile of 275 counts.
-Therefore, the unevenness in the simulated data reasonably reflects that of a real experiment, and so these data already represents a fair test for the inferences.
+Therefore, the unevenness in the simulated data reasonably reflects that of a real experiment, and so these data already represent a fair test for the inferences.
 

### 2015-03-24 paper/BMC_Bioinformatics_2015/bmc_bioinformatics_article.tex
Final check of reviewer responses

@@
 The program \texttt{dms\_logoplot} visualizes the posterior mean differential preferences via an extension to \texttt{weblogo}~\cite{crooks2004}.
-In addition, \texttt{dms\_inferdiffprefs} also creates text files that give the posterior probability that $\Delta\pi_{r,x} > 0$ or $< 0$ (these posterior probabilities are \emph{not} corrected to account for the fact that multiple sites are typically being examined, although be default the inferences are made using the regularizing prior in Equation \ref{eq:pr_deltapi}).
+In addition, \texttt{dms\_inferdiffprefs} creates text files that give the posterior probability that $\Delta\pi_{r,x} > 0$ or $< 0$.
+These posterior probabilities are \emph{not} corrected to account for the fact that multiple sites are typically being examined, although be default the inferences are made using the regularizing prior in Equation \ref{eq:pr_deltapi}.
 

### 2015-03-24 paper/BMC_Bioinformatics_2015/bmc_bioinformatics_article.tex

@@
 I describe a software package, \texttt{dms\_tools}, to infer the impacts of mutations from deep mutational scanning data using a likelihood-based treatment of the mutation counts.
-I show that \texttt{dms\_tools} yields more accurate inferences on simulated data than the widely used but statistically biased approach of calculating ratios of counts pre- and post-selection.
+I show that \texttt{dms\_tools} yields more accurate inferences on simulated data than simply calculating ratios of counts pre- and post-selection.
 Using \texttt{dms\_tools}, one can infer the preference of each site for each amino acid given a single selection pressure, or assess the extent to which these preferences change under different selection pressures.
@@
 Since the original description of deep mutational scanning by Fowler \textit{et al}~\cite{fowler2010high}, the technique has been applied to a wide range of genes~\cite{traxlmayr2012construction,mclaughlin2012spatial,starita2013activity,melamed2013deep,roscoe2013analyses,firnberg2014comprehensive,bloom2014experimentally,melnikov2014comprehensive,thyagarajan2014inherent,wu2014high,wu2014ns,olson2014comprehensive,kitzman2015massively}, both to measure mutational tolerance given a single selection pressure as in Figure \ref{fig:DeepMutationalScan}A, or to identify mutations that have different effects under alternative selections as in Figure \ref{fig:DeepMutationalScan}B.
-New techniques to create comprehensive codon-mutant libraries of genes make it possible to profile all single amino-acid mutations~\cite{firnberg2012,jain2013rapid,firnberg2014comprehensive,bloom2014experimentally,melnikov2014comprehensive,kitzman2015massively}, while new techniques for targeted mutagenesis of mammalian genomes enable deep mutational scanning to be applied across the biological spectrum from viruses and bacteria to human cells~\cite{findlay2014saturation}.
+New techniques to create comprehensive codon-mutant libraries of genes make it possible to profile all amino-acid mutations~\cite{firnberg2012,jain2013rapid,firnberg2014comprehensive,bloom2014experimentally,melnikov2014comprehensive,kitzman2015massively}, while new techniques for targeted mutagenesis of mammalian genomes enable deep mutational scanning to be applied across the biological spectrum from viruses and bacteria to human cells~\cite{findlay2014saturation}.
 
@@
 Properly done experiments should quantify the number of unique molecules in the library so that it is obvious whether this assumption holds.
-In the absence of such information, the analysis can be repeated using only a random fraction of the deep sequencing data to assess whether inferences are limited by sequencing depth or they underlying molecular diversity in the mutant library.
+In the absence of such information, the analysis can be repeated using only a random fraction of the deep sequencing data to assess whether inferences are limited by sequencing depth or the underlying molecular diversity in the mutant library.
 
@@
 The program \texttt{dms\_inferprefs} in the \texttt{dms\_tools} package infers the preferences by using \texttt{pystan}~\cite{pystan-software:2014} to perform MCMC over the posterior defined by the product of the likelihoods and priors in Equations \ref{eq:pr_nrerrpre}, \ref{eq:pr_nrerrpost}, \ref{eq:pr_nrpre}, \ref{eq:pr_nrpost}, \ref{eq:pr_pir}, \ref{eq:pr_mur}, \ref{eq:pr_epsilonr}, and \ref{eq:pr_rhor}.
-The program runs four chains from different initial values, and checks for convergence by ensuring that the Gelman-Rubin statistic $\hat{R}$~\cite{gelman1992inference} is $<1.1$ and the effective sample size is $>100$; the number of MCMC iterations is increased until these convergence is achieved.
+The program runs four chains from different initial values, and checks for convergence by ensuring that the Gelman-Rubin statistic $\hat{R}$~\cite{gelman1992inference} is $<1.1$ and the effective sample size is $>100$; the number of MCMC iterations is increased until convergence is achieved.
 The program \texttt{dms\_logoplot} in the \texttt{dms\_tools} package visualizes the posterior mean preferences via an extension to \texttt{weblogo}~\cite{crooks2004}.
@@
 Figures \ref{fig:MelnikovPrefs} and \ref{fig:ThyagarajanPrefs} also illustrate use of \texttt{dms\_correlate} to assess the correlation between preferences inferred from different biological replicates~\cite{blainey2014points} of the experiment.
-The inclusion and analysis of such replicates provide the only sure way to fully assess the sources of noise associated with deep mutational scanning.
+The inclusion and analysis of such replicates is the only sure way to fully assess the sources of noise associated with deep mutational scanning.
 
@@
 \subsubsection*{Is the Bayesian inference worthwhile?}
-The foregoing sections explain why the Bayesian inference of preferences implemented in \texttt{dms\_tools} is conceptually preferable to estimating mutational effects via direct calculation of enrichment ratios using Equation \ref{eq:enrichmentcounts}.
+The foregoing sections explain why the Bayesian inference of preferences implemented in \texttt{dms\_tools} is conceptually preferable to estimating mutational effects via direct ratio estimation using Equation \ref{eq:enrichmentcounts}.
 However, do the practical benefits of this Bayesian inference justify its increased complexity?
-The simulations in the previous section show that the Bayesian inference is more accurate, but in the absence of background errors (Figure \ref{fig:PrefCorrelations}A) the magnitude of the improvement becomes negligible once the mutations counts per site $N \overline{\mu}$ start to exceed $\sim10^3$.
+The simulations in the previous section show that the Bayesian inference is more accurate, but in the absence of background errors (Figure \ref{fig:PrefCorrelations}A) the magnitude of the improvement becomes negligible once the mutation counts per site $N \overline{\mu}$ start to exceed $\sim10^3$.
 When there is a need to correct for background errors (Figure \ref{fig:PrefCorrelations}B), meaningful benefits of the Bayesian inference over enrichment ratios extend to somewhat higher sequencing depths.
-Overall, it appears that Bayesian inference will always perform as well or better than enrichment ratios.
-But the tangible benefit becomes negligible at high sequencing depth.
+Overall, it appears that Bayesian inference will always perform as well or better than ratio estimation, but that the tangible benefit becomes negligible at high sequencing depth.
 In that case, the user will have to decide if the increased computational runtime and complexity of the Bayesian inference is worth a small improvement in accuracy.
-Simpler enrichment-ratio calculations can be performed using the \texttt{--ratio\_estimation} option of \texttt{dms\_inferprefs} or using an alternative program such as \texttt{Enrich}~\cite{fowler2011enrich}.
-When applying enrichment-ratio calculations to data where some mutations have low counts, it is important to include pseudocounts (denoted by $\mathcal{P}$ in Equation \ref{eq:enrichmentcounts}) as a form of regularization to avoid estimating excessively high or low ratios at sites with limited counts.
+Simpler ratio estimation can be performed using the \texttt{--ratio\_estimation} option of \texttt{dms\_inferprefs} or using an alternative program such as \texttt{Enrich}~\cite{fowler2011enrich}.
+When applying ratio estimation to data where some mutations have low counts, it is important to include pseudocounts (denoted by $\mathcal{P}$ in Equation \ref{eq:enrichmentcounts}) as a form of regularization to avoid estimating excessively high or low preferences at sites with limited counts.
 
@@
 
-But in Figure \ref{fig:DeepMutationalScan}B, the assumption that most mutations will be similarly selected is a plausible null hypothesis, since we expect to alternative selections to have markedly different effects on only a small subset of mutations (typically, major constraints related to protein folding and stability will be relatively conserved across different selections on the same protein).
+But in Figure \ref{fig:DeepMutationalScan}B, the assumption that most mutations will be similarly selected is a plausible null hypothesis, since we expect alternative selections to have markedly different effects on only a small subset of mutations (typically, major constraints related to protein folding and stability will be conserved across different selections on the same protein).
 Therefore, \texttt{dms\_tools} uses a different algorithm to infer the differential preferences under the two selections.
@@
 Equations \ref{eq:controlpi} and \ref{eq:deltapi} imply
-$\boldsymbol{\mathbf{f_r^{s1}}}= \frac{\boldsymbol{\mathbf{f_r^{\textbf{starat}}}}\circ \boldsymbol{\mathbf{\pi_r^{s1}}}}{\boldsymbol{\mathbf{f_r^{\textbf{start}}}}\cdot \boldsymbol{\mathbf{\pi_r^{s1}}}}$ and $\boldsymbol{\mathbf{f_r^{s2}}} = \frac{\boldsymbol{\mathbf{f_r^{\textbf{start}}}}\circ \left(\boldsymbol{\mathbf{\pi_r^{s1}}} + \boldsymbol{\mathbf{\Delta\pi_r}}\right)}{\boldsymbol{\mathbf{f_r^{\textbf{start}}}}\cdot \left(\boldsymbol{\mathbf{\pi_r^{s1}}} + \boldsymbol{\mathbf{\Delta\pi_r}}\right)}$.
+$\boldsymbol{\mathbf{f_r^{s1}}}= \frac{\boldsymbol{\mathbf{f_r^{\textbf{start}}}}\circ \boldsymbol{\mathbf{\pi_r^{s1}}}}{\boldsymbol{\mathbf{f_r^{\textbf{start}}}}\cdot \boldsymbol{\mathbf{\pi_r^{s1}}}}$ and $\boldsymbol{\mathbf{f_r^{s2}}} = \frac{\boldsymbol{\mathbf{f_r^{\textbf{start}}}}\circ \left(\boldsymbol{\mathbf{\pi_r^{s1}}} + \boldsymbol{\mathbf{\Delta\pi_r}}\right)}{\boldsymbol{\mathbf{f_r^{\textbf{start}}}}\cdot \left(\boldsymbol{\mathbf{\pi_r^{s1}}} + \boldsymbol{\mathbf{\Delta\pi_r}}\right)}$.
 
@@
 In addition, \texttt{dms\_inferdiffprefs} creates text files that give the posterior probability that $\Delta\pi_{r,x} > 0$ or $< 0$.
-These posterior probabilities are \emph{not} corrected to account for the fact that multiple sites are typically being examined, although be default the inferences are made using the regularizing prior in Equation \ref{eq:pr_deltapi}.
+These posterior probabilities are \emph{not} corrected to account for the fact that multiple sites are typically being examined, although by default the inferences are made using the regularizing prior in Equation \ref{eq:pr_deltapi}.
 
@@
 \section*{Data and code for figures in this paper.}
-The data and computer code used to generate the figures are in the tagged version of the \texttt{dms\_tools} source code at \url{https://github.com/jbloom/dms_tools} in \texttt{examples} subdirectory.
+The data and computer code used to generate the figures are in version 1.01 of the \texttt{dms\_tools} source code (which is tagged on \texttt{Github} at \url{https://github.com/jbloom/dms_tools/tree/1.0.1}) in the \texttt{examples} subdirectory.
 The \texttt{LaTex} source for this paper is in the \texttt{paper} subdirectory.

### 2015-03-24 paper/BMC_Bioinformatics_2015/bmc_bioinformatics_article_nofigs.tex

@@
 Deep sequencing of the variants pre- and post-selection provides information about the functional impacts of mutations.
-Since the original description of deep mutational scanning by Fowler \textit{et al}~\cite{fowler2010high}, the technique has been applied to a wide range of genes~\cite{traxlmayr2012construction,mclaughlin2012spatial,starita2013activity,melamed2013deep,roscoe2013analyses,firnberg2014comprehensive,bloom2014experimentally,melnikov2014comprehensive,thyagarajan2014inherent,wu2014high,wu2014ns,olson2014comprehensive}, both to measure mutational tolerance given a single selection pressure as in Figure \ref{fig:DeepMutationalScan}A, or to identify mutations that have different effects under alternative selections as in Figure \ref{fig:DeepMutationalScan}B.
-New techniques to create comprehensive codon-mutant libraries of genes make it possible to profile all single amino-acid mutations~\cite{firnberg2012,jain2013rapid,firnberg2014comprehensive,bloom2014experimentally,melnikov2014comprehensive}, while new techniques for targeted mutagenesis of mammalian genomes enable deep mutational scanning to be applied across the biological spectrum from viruses and bacteria to human cells~\cite{findlay2014saturation}.
+Since the original description of deep mutational scanning by Fowler \textit{et al}~\cite{fowler2010high}, the technique has been applied to a wide range of genes~\cite{traxlmayr2012construction,mclaughlin2012spatial,starita2013activity,melamed2013deep,roscoe2013analyses,firnberg2014comprehensive,bloom2014experimentally,melnikov2014comprehensive,thyagarajan2014inherent,wu2014high,wu2014ns,olson2014comprehensive,kitzman2015massively}, both to measure mutational tolerance given a single selection pressure as in Figure \ref{fig:DeepMutationalScan}A, or to identify mutations that have different effects under alternative selections as in Figure \ref{fig:DeepMutationalScan}B.
+New techniques to create comprehensive codon-mutant libraries of genes make it possible to profile all amino-acid mutations~\cite{firnberg2012,jain2013rapid,firnberg2014comprehensive,bloom2014experimentally,melnikov2014comprehensive,kitzman2015massively}, while new techniques for targeted mutagenesis of mammalian genomes enable deep mutational scanning to be applied across the biological spectrum from viruses and bacteria to human cells~\cite{findlay2014saturation}.
 
@@
 
+The rest of this paper assumes that the sequencing depth is less than the number of unique molecules in the mutant library, such that the deep sequencing randomly subsamples the set of molecules.
+If this assumption is false (i.e.
+if the number of unique molecules is substantially less than the sequencing depth), then the accuracy of inferences about mutational effects will be fundamentally limited by this aspect of the experimental design.
+Properly done experiments should quantify the number of unique molecules in the library so that it is obvious whether this assumption holds.
+In the absence of such information, the analysis can be repeated using only a random fraction of the deep sequencing data to assess whether inferences are limited by sequencing depth or the underlying molecular diversity in the mutant library.
+
 \subsection*{The goal: inferring site-specific amino-acid preferences.}
 The goal is to estimate the effects of mutations from changes in their counts after selection.
-Let $\mu_{r,x}$, $f_{r,x}$, $f_{r,x}^{s1}$, and $f_{r,x}^{s2}$ denote the \emph{true} frequencies of $x$ at $r$ that would be observed for the four libraries in Figure \ref{fig:DeepMutationalScan} if we sampled at infinite depth in both the actual experiment and the sequencing.
-If we knew these frequencies, we could calculate parameters that reflect the effects of mutations.
+Let $\mu_{r,x}$, $f_{r,x}$, $f_{r,x}^{s1}$, and $f_{r,x}^{s2}$ denote the \emph{true} frequencies at site $r$ of all mutant characters $x \ne \operatorname{wt}\left(r\right)$ that would be observed for the four libraries in Figure \ref{fig:DeepMutationalScan} if we sampled at infinite depth in both the actual experiment and the sequencing.
+The definition of these frequencies for the wildtype character $\operatorname{wt}\left(r\right)$ depends on how the mutant library is constructed.
+If the mutant library is constructed so that there is a Poisson distribution of the number of mutations per gene (as is the case for error-prone PCR or the codon-mutagenesis in \cite{bloom2014experimentally,thyagarajan2014inherent}), then $\mu_{r,\operatorname{wt}\left(r\right)}$, $f_{r,\operatorname{wt}\left(r\right)}$, $f_{r,\operatorname{wt}\left(r\right)}^{s1}$, and $f_{r,\operatorname{wt}\left(r\right)}^{s2}$ are defined as for all other characters $x$, and denote the frequencies of $\operatorname{wt}\left(r\right)$ at site $r$ that would be observed if sampling at infinite depth.
+The reason we can make this definition for libraries containing genes with Poisson-distributed numbers of mutations is that for any reasonable-length gene ($L \gg 1$), the marginal distribution of the number of mutations in a gene is virtually unchanged by the knowledge that there is a mutation at site $r$.
+On the other hand, if the mutant library is constructed so that there is exactly zero or one mutation per gene (as in \cite{firnberg2014comprehensive,melnikov2014comprehensive,kitzman2015massively}), then the marginal distribution of the total number of mutations in a gene is changed by the knowledge that there is a mutation at $r$.
+In this case, the wildtype-character frequencies $\mu_{r,\operatorname{wt}\left(r\right)}$, $f_{r,\operatorname{wt}\left(r\right)}$, $f_{r,\operatorname{wt}\left(r\right)}^{s1}$, and $f_{r,\operatorname{wt}\left(r\right)}^{s2}$ are correctly defined as the frequency of unmutated genes in the library, and the counts $n_{r,\operatorname{wt}\left(r\right)}^{\rm{pre}}$, etc.
+are defined as the number of reads at $r$ attributable to unmutated genes.
+In this case, measurement of these counts requires sequencing with linkage as in \cite{hiatt2010parallel,wu2014hiv,kitzman2015massively}.
+The proper analysis of libraries containing only unmutated and singly mutated clones sequenced without linkage is beyond the scope of this paper.
+
+If we knew the frequencies $\mu_{r,x}$, $f_{r,x}$, $f_{r,x}^{s1}$, and $f_{r,x}^{s2}$, we could calculate parameters that reflect the effects of mutations.
 One parameter that characterizes the effect of mutating $r$ from $\operatorname{wt}\left(r\right)$ to $x$ for the experiment in Figure \ref{fig:DeepMutationalScan}A is the \emph{enrichment ratio}, which is the relative frequency of mutations to $x$ after selection versus before selection:
@@
 The program \texttt{dms\_inferprefs} in the \texttt{dms\_tools} package infers the preferences by using \texttt{pystan}~\cite{pystan-software:2014} to perform MCMC over the posterior defined by the product of the likelihoods and priors in Equations \ref{eq:pr_nrerrpre}, \ref{eq:pr_nrerrpost}, \ref{eq:pr_nrpre}, \ref{eq:pr_nrpost}, \ref{eq:pr_pir}, \ref{eq:pr_mur}, \ref{eq:pr_epsilonr}, and \ref{eq:pr_rhor}.
-The program runs four chains from different initial values, and checks for convergence by ensuring that the Gelman-Rubin statistic $\hat{R}$~\cite{gelman1992inference} is $<1.1$ and the effective sample size is $>100$; the number of MCMC iterations is increased until these convergence is achieved.
+The program runs four chains from different initial values, and checks for convergence by ensuring that the Gelman-Rubin statistic $\hat{R}$~\cite{gelman1992inference} is $<1.1$ and the effective sample size is $>100$; the number of MCMC iterations is increased until convergence is achieved.
 The program \texttt{dms\_logoplot} in the \texttt{dms\_tools} package visualizes the posterior mean preferences via an extension to \texttt{weblogo}~\cite{crooks2004}.
+The program \texttt{dms\_merge} can be used to average preferences inferred from different experimental replicates that have individually been analyzed by \texttt{dms\_inferprefs}, and the program \texttt{dms\_correlate} can be used to compute the correlations among inferences from different replicates.
 
@@
 Inferences made by \texttt{dms\_tools} are always more accurate than those obtained by simply taking ratios of mutation counts.
+
+\subsubsection*{Is the Bayesian inference worthwhile?}
+The foregoing sections explain why the Bayesian inference of preferences implemented in \texttt{dms\_tools} is conceptually preferable to estimating mutational effects via direct ratio estimation using Equation \ref{eq:enrichmentcounts}.
+However, do the practical benefits of this Bayesian inference justify its increased complexity?
+The simulations in the previous section show that the Bayesian inference is more accurate, but in the absence of background errors (Figure \ref{fig:PrefCorrelations}A) the magnitude of the improvement becomes negligible once the mutation counts per site $N \overline{\mu}$ start to exceed $\sim10^3$.
+When there is a need to correct for background errors (Figure \ref{fig:PrefCorrelations}B), meaningful benefits of the Bayesian inference over enrichment ratios extend to somewhat higher sequencing depths.
+Overall, it appears that Bayesian inference will always perform as well or better than ratio estimation, but that the tangible benefit becomes negligible at high sequencing depth.
+In that case, the user will have to decide if the increased computational runtime and complexity of the Bayesian inference is worth a small improvement in accuracy.
+Simpler ratio estimation can be performed using the \texttt{--ratio\_estimation} option of \texttt{dms\_inferprefs} or using an alternative program such as \texttt{Enrich}~\cite{fowler2011enrich}.
+When applying ratio estimation to data where some mutations have low counts, it is important to include pseudocounts (denoted by $\mathcal{P}$ in Equation \ref{eq:enrichmentcounts}) as a form of regularization to avoid estimating excessively high or low preferences at sites with limited counts.
 
@@
 The program \texttt{dms\_logoplot} visualizes the posterior mean differential preferences via an extension to \texttt{weblogo}~\cite{crooks2004}.
-In addition, \texttt{dms\_inferdiffprefs} also creates text files that give the posterior probability that $\Delta\pi_{r,x} > 0$ or $< 0$ (in other words, the probability that there is differential selection).
+In addition, \texttt{dms\_inferdiffprefs} creates text files that give the posterior probability that $\Delta\pi_{r,x} > 0$ or $< 0$.
+These posterior probabilities are \emph{not} corrected to account for the fact that multiple sites are typically being examined, although by default the inferences are made using the regularizing prior in Equation \ref{eq:pr_deltapi}.
 
@@
 %\centerline{\includegraphics[width=6.5in]{../../examples/Melnikov_et_al_Tn5/infer_prefs_on_simulated_data/correlations.pdf}}
-\cprotect\caption{\label{fig:PrefCorrelations}\csentence{Accuracy of preference inference on simulated data.} Deep mutational scanning counts were simulated using the preferences in Figure \ref{fig:MelnikovPrefs}A and realistic parameters for mutation and error rates.
+\cprotect\caption{\label{fig:PrefCorrelations}\csentence{Accuracy of preference inference on simulated data.} Deep mutational scanning counts were simulated using the preferences in Figure \ref{fig:MelnikovPrefs}A and realistic mutation and error rates that were uneven across sites and characters as in actual experiments.
 The simulations were done {\bf (A)} without or {\bf (B)} with sequencing errors quantified by control libraries.
